---
title: "DataPreparation"
author: "Aimara Planillo"
date: "1/28/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare data for Wolf Habitat Suitability - GERMANY 

### Load packages and workspace from source file
```{r}
source(paste0(getwd(), "/R/source_file.R"))
workDir <- getwd()
dataproc_wd <- file.path(workDir, "output/data-proc")

set.seed(155)
```


####################
## telemetry data ##
####################

This is already processed data provided by Moritz. Data was selected to represent only 
resident adult individuals, which are the ones used in the models.

## Load and check telemetry
```{r}
# load all GPS data for resident adults
gps_points <- readRDS("data-raw/gps_points_3035_crop_adults_20190130.rds")
table(gps_points$status)
nrow(gps_points)
# [1] 46304
length(unique(gps_points$animalID))
# [1] 20
table(gps_points$animalID)

# basic summaries
gps_summary <- gps_points %>%
  dplyr::select(-geometry) %>% 
  group_by(animalID) %>%
  summarise(n = n()) 
mean(gps_summary$n)
# [1] 2315.2 # average number of points per individual
sd(gps_summary$n)
# [1] 2153.612 # sd of the number of points per ind
range(gps_summary$n)
# [1]  142 8499 # min and max values of the number of points per ind


##  Plot the locations colorcode by individual
tmap_mode("view")
tm_shape(as_Spatial(gps_points)) +
  tm_dots("animalID")

# check how the points per individual look like
## Number of points per territory. Individuals with no named territories show 0 values
table(gps_points$territory, gps_points$animalID)

# Individual ID1 Has two territories, we divide them based on the name of territory 
# into two "effective individuals". 

gps_21ter <- gps_points %>%
  mutate(id = case_when(
    animalID == "ID1" & territory == "Prösa" ~ "ID1p", 
    TRUE ~ animalID))

table(gps_21ter$id)

# check on map it territories look correct
tm_shape(as_Spatial(gps_21ter)) +
  tm_dots("id")

# check if the temporal line for both territories in ID1 makes sense
tmap_mode("plot")
gps_21ter %>%
  filter(animalID == "ID1") %>%
  mutate(order = seq(1, nrow(gps_21ter[gps_21ter$animalID == "ID1",]))) %>% 
  group_by(id) %>% 
  as_Spatial() %>% 
  tm_shape() +
  tm_dots(col = "order", shape = "id", size = 1)

gps_21ter %>%
   filter(animalID == "ID1") %>%
   ggplot(aes(timestamp, id)) +
   geom_point(aes(col = territory)) +
   scale_x_datetime(labels = scales::date_format("%m.%Y"), 
                    breaks = seq(ISOdate(2009,1,1), 
                                 by = "3 months", 
                                 length.out = 40))  
    
# Some points seem to overlap a bit around year 2018, more in detail:
test <- gps_21ter %>% 
  filter(animalID == "ID1") %>% 
  mutate(year = case_when(
    timestamp >= "2018-01-01" ~ "af2018",
    TRUE ~ "bf2018")) %>% 
  group_by(id, year) %>% 
  summarise(territory_id = "territory",
            n = n()) 
test            

## Remove the "overlapping points as transient phase
gps_21ter2 <- gps_21ter %>%
  mutate(year = case_when(
    timestamp >= "2018-01-01" ~ "af2018",
    TRUE ~ "bf2018")) %>% 
  filter(!(id == "ID1" & year == "af2018")) %>%
  filter(!(id == "ID1p" & year == "bf2018"))
nrow(gps_21ter2)
# [1] 46175 # 129 gps locations removed
```

## Create MCPs

We are going to create the 95% MCPs from the gps locations to extract home range information.
Then, we select data before 2018 for running the model and create a 100% MCP for background points in the models

```{r}
## 95% MCPs

# Only include three columns (id, x, and y coordinates) for making MCP's
mcp_dataframe <- gps_21ter2 %>%
  as.data.frame() %>%
  dplyr::select(id) 
mcp_dataframe <- cbind(mcp_dataframe, st_coordinates(gps_21ter2))
head(mcp_dataframe)
nrow(mcp_dataframe)
# [1] 46175 # Same as gps locations

# To compute the mcp in adehabitatHR package we need an sp object
#Create a SpatialPointsDataFrame by defining the coordinates
coordinates(mcp_dataframe) <- c("X", "Y")
# Set the coordinate reference system (CRS)
proj4string(mcp_dataframe) <- CRS( "+init=epsg:3035" )

# Calculate 95% MCPs for each territory
mcp95_all <- mcp(mcp_dataframe, percent = 95)

# Examine output
mcp95_all
plot(mcp95_all)

nrow(mcp95_all)
# [1] 21 # Number of territories. Remember that id1 was split
length(unique(mcp95_all$id))

mcp95_all$color <- as.character(viridis(21)) #add a column to give colours to the polygons based on viridis palette
tm_shape(mcp95_all) +
  tm_borders(col = mcp95_all$color) 

# save 95% mcp as geopackage
mcp95_all_sf <- st_as_sf(mcp95_all, crs = 3035)
plot(st_geometry(mcp95_all_sf))

## Get starting and ending date of data for each individual
gps_dates <- gps_21ter2 %>% 
  group_by(id) %>% 
  arrange(timestamp) %>% 
  summarise(start_date = first(timestamp),
            end_date = last(timestamp)) %>% 
  as.data.frame() %>% 
  dplyr::select(-geometry)

## Add dates to mcp 
mcp95_all_dates <- left_join(gps_dates, mcp95_all_sf, by = "id") %>% 
  st_as_sf
# write_sf(mcp95_all_sf, paste0(dataproc_wd, "/gps_MCP95_adults_allpoints_3035.gpkg"))


## 100% MCPs before 2018

# We divide dataset before and after 2018
gps_bf2018 <- gps_21ter2 %>%
  filter(year == "bf2018")
length(unique(gps_bf2018$id))
#  18

gps_af2018 <- gps_21ter2 %>%
  filter(year == "af2018")  
length(unique(gps_af2018$id))
# 5

## Calculate 100 MCPs for each individual before 2018
mcp_bf2018_dataframe <- gps_bf2018 %>%
  as.data.frame() %>%
  dplyr::select(id) %>% 
  droplevels()
mcp_bf2018_dataframe <- cbind(mcp_bf2018_dataframe, st_coordinates(gps_bf2018))
head(mcp_bf2018_dataframe)
nrow(mcp_bf2018_dataframe)
# [1] 28122 # locations before 2018
table(mcp_bf2018_dataframe$id)

coordinates(mcp_bf2018_dataframe) <- c("X", "Y")
proj4string(mcp_bf2018_dataframe) <- CRS( "+init=epsg:3035" )

mcp100_bf2018 <- mcp(mcp_bf2018_dataframe, percent = 100)
mcp100_bf2018
nrow(mcp100_bf2018)
# [1] 18
gps_mcp100_bf2018 <- st_as_sf(mcp100_bf2018, crs = 3035)
plot(st_geometry(gps_mcp100_bf2018))
# write_sf(gps_mcp100_bf2018, paste0(dataproc_wd, "/gps_MCP100_adults_bf2018_3035.gpkg"))
```

Get estimates of area from the MCPs
```{r}
# area of MCPs:
mcp95_km2 <- st_area(mcp95_all_sf)/1000000

## Average values of the 95% MCPs
mcp95_km2 %>% 
  as.data.frame() %>% 
  summarise_all(list(mean = mean, sd = sd, 
                                      median = median, 
                                      min = min, max = max)) %>% 
  t()
#            [,1]
# mean   234.4418
# sd     129.0340
# median 244.6626
# min     45.8442
# max    461.7623
```


## Filter telemetry data as 1 observation per day and 1 per grid cell 
We do this to avoid temp and spatial autocorrelation

A. Select one observation per day per individual
```{r}
gps21_summary <- gps_21ter2 %>%
  group_by(id) %>%
  summarise(n = n())

length(unique(gps_21ter2$id))
nrow(gps21_summary)
mean(gps21_summary$n)
# [1] 2198.81 # average points per ind
sd(gps21_summary$n)
# [1] 2140.341 # sd points per ind
range(gps21_summary$n)
# [1]  142 8499 # min max points per ind:

gps21_summary$id[gps21_summary$n == 142]
# [1] "MT3"
gps21_summary$id[gps21_summary$n == 8499]
# [1] "WR4"

# temporal scale of data
head(gps_21ter2)
times <- gps_21ter2$timestamp 
# differences in hours bt observatinons
times_dif <- difftime(times[-length(times)], times[-1], unit = "hour") 
times_dif <- c(0, as.numeric(times_dif)) # we add 0 to the first difference
times_dif <- cbind.data.frame(id = gps_21ter2$id, times_dif)
str(times_dif)
summary(times_dif)

# to calculate average time, we remove first entry because it does not make sense
times_dif <- times_dif[duplicated(times_dif$id),]

# now we remove values that are too high 
times_dif_filtered <- times_dif %>% 
  filter(times_dif %in% (-30:30)) # keep values between -30 and 30 hours

times_dif[times_dif$id == "ID1",]

summary_times_dif <- times_dif_filtered %>%
  group_by(id) %>%
  summarise(mean = mean(times_dif), 
            sd = sd(times_dif))

as.data.frame(summary_times_dif)

# Usually 1 hour or 4 hours intervals. We check weird numbers:
gps_21ter2[gps_21ter2$id == "FT10",] # 1h interval
gps_21ter2[gps_21ter2$id == "FT9",] # 4h interval
gps_21ter2[gps_21ter2$id == "MT3",] # 4h interval
as.data.frame(gps_21ter2[gps_21ter2$id == "WF3", ]) # Irregular: 1,2,3,6 h interval
gps_21ter2[gps_21ter2$id == "WR1",] # 3h interval
as.data.frame(gps_21ter2[gps_21ter2$id == "WF5", ]) # 1-3h interval


# extract date for each observation, group by date and individual 
#(using the newly created id column to separate the two territories for individual id1!!)
#and select a random observation
gps_points_filter_day <- gps_21ter2 %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(id, date) %>%
  sample_n(1)

nrow(gps_points_filter_day)
# [1] 4949

# plot to check
tm_shape(gps_21ter2) + # all gps points in red
  tm_dots("red") +
  tm_shape(gps_points_filter_day) + #filtered gps points in blue
  tm_dots("blue") 
# there is a clear difference in the number of points, as expected. 

table(gps_points_filter_day$id)
```


B. Select one observation per cell 

```{r}
# load land cover raster to use the resolution as reference for the grids
envir_stack <- stack("./data-raw/geo-raw/stack_all_env_vars_20181220.tif")
names(envir_stack) <- c("g100_clc12_V18_5", "popu01clcv5", "mos2_dist_settlments", "mos2_dist_streets", "HFP2009_int")
copernicus_lc_3035 <- envir_stack$g100_clc12_V18_5

# extract coordinates of 1 observation per grid cell
gps_filtered_day_grid <- gridSample(st_coordinates(gps_points_filter_day), 
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() 

nrow(gps_filtered_day_grid)
# [1] 3841

# transform to sf object
gps_filtered_day_grid <- st_as_sf(gps_filtered_day_grid, 
                            coords = c("X","Y"), 
                            crs = 3035, 
                            sf_column_name = "geometry") # we need this parameter to run the function

# add the individual Id by spatial join
gps_filtered_day_grid_ind <-  st_join(gps_filtered_day_grid, gps_21ter2, 
                                      join = st_nearest_feature,
                                      left = TRUE)
nrow(gps_filtered_day_grid_ind)
# [1] 3841

colnames(gps_filtered_day_grid_ind)
gps_filtered_day_grid_ind %>% 
  group_by(id) %>%
  summarise(gps_id = n()) %>%
  summarise(min = min(gps_id),
            max = max(gps_id), 
            mean = mean(gps_id), 
            sd = sd(gps_id)) %>% 
  as.data.frame()  
# min max     mean       sd              
# 23 418 182.9048 104.2257 
  
table(gps_filtered_day_grid_ind$id)
```


## Divide into evaluation, training and test sets

```{r}
# Select points after 2018 for evaluation
gps_filter_af2018 <- gps_filtered_day_grid_ind %>% 
  filter(year == "af2018") 

# write_sf(gps_filter_af2018, paste0(dataproc_wd, "/gps_points_eval_af2018_3035.gpkg"))

# Select points before 2018 for train and test sets
gps_bf2018 <- gps_filtered_day_grid_ind %>% 
  filter(year == "bf2018") %>% 
  as.data.frame() %>% 
  droplevels()
nrow(gps_bf2018)
# [1] 3240

# divide by individual id
id_list <- split(gps_bf2018, gps_bf2018$id)
id_list[1][[1]]

# divide 4:1 points of each id
for (i in 1:length(id_list)){
  id_list[i][[1]]$fold <- kfold(id_list[i][[1]], k = 5)
}

gps_bf20182 <- do.call(rbind.data.frame, id_list)
gps_bf20182 <- gps_bf20182[,c(1:(ncol(gps_bf20182)-2), ncol(gps_bf20182), ncol(gps_bf20182)-1)] # reorders and puts the fold in the column before the last one
nrow(gps_bf20182)
# [1] 3240

gps_train <- gps_bf20182[gps_bf20182$fold!=5,] 
gps_test <- gps_bf20182[gps_bf20182$fold==5,]

nrow(gps_train)
# [1] 2591
nrow(gps_test)
# [1] 649

# plot train and test data
pres_train_sf <- st_sf(gps_train)
pres_test_sf <- st_sf(gps_test) 

tmap_mode("view")
tm_shape(pres_train_sf) + #filtered training gps points in color per id 
  tm_dots("id") +
  tm_shape(pres_test_sf) + #filtered testing gps points in light blue
  tm_dots("lightblue")

# write_sf(pres_train_sf,
#         paste0(dataproc_wd,"/gps_points_train_bf2018_3035.gpkg"))
# write_sf(pres_test_sf,
#         paste0(dataproc_wd,"/gps_points_test_bf2018_3035.gpkg"))
```


## Background points bf 2018 for GLMM from MCPs

Extract same number of bg points as presences and divide into training and testing sets
```{r}
# Select points before 2018 for train and test sets
gps_bf2018 <- gps_filtered_day_grid_ind %>% 
  filter(year == "bf2018") %>% 
  as.data.frame() %>% 
  droplevels()
nrow(gps_bf2018)

# add number of points to mcp object
mcp100_bf2018_count <- gps_bf2018 %>% 
  group_by(id) %>% 
  summarise(n = n()) %>% 
  as.data.frame() %>% 
  dplyr::select(id, n) %>% 
  ungroup() %>% 
  right_join(gps_mcp100_bf2018, by = "id") %>% 
  st_sf() %>% 
  as_Spatial() # we need it as spatial object for the next function

plot(mcp100_bf2018_count)

nrow(mcp100_bf2018_count)
# 18

# Extract 10 times same amount of background points as presences from MCPs
uni_id <- mcp100_bf2018_count$id
bg_points_tmp <- list()

for(i in 1:length(uni_id)){
  bg_points_tmp[[i]] <- spsample(mcp100_bf2018_count[i,], n = as.numeric(mcp100_bf2018_count$n[i])*10, type = "random") %>%
    as.data.frame()
  bg_points_tmp[[i]]$id <- mcp100_bf2018_count$id[i] # add territory id
  names(bg_points_tmp)[i] <- uni_id[i]
}

names(bg_points_tmp)

bg_points_tmp2 <- as.data.frame(do.call(rbind, bg_points_tmp))

# add same color as territories
bg_points_tmp2 <- bg_points_tmp2 %>% 
  left_join(as.data.frame(mcp95_all_sf)[,c(1,3)], by = "id") %>% 
  droplevels() # remove id with 0 observations
head(bg_points_tmp2)

# divide again by id
bg_id_list <- split(bg_points_tmp2, bg_points_tmp2$id)
bg_id_list[1][[1]]
str(bg_id_list)
length(bg_id_list)
# [1] 18

# divide 4:1 points of each id
for (i in 1:length(bg_id_list)){
  bg_id_list[i][[1]]$fold <- kfold(bg_id_list[i][[1]], k = 5)
}


tmp2 <- do.call(rbind.data.frame, bg_id_list)
nrow(tmp2)
# [1] 32400
with(tmp2, table(color, id))
summary(tmp2)

bg_train_glmm <- tmp2[tmp2$fold!=5,] 
bg_test_glmm <- tmp2[tmp2$fold==5,]

nrow(bg_train_glmm)
# [1] 25920
nrow(bg_test_glmm)
# [1] 6480

# make sf objects
bg_train_glmm_sf <- st_as_sf(bg_train_glmm, coords = c("x", "y"), crs = 3035) 
bg_test_glmm_sf <- st_as_sf(bg_test_glmm, coords = c("x", "y"), crs = 3035) 


# select one observation per grid cell
# extract coordinates of 1 observation per grid cell
bg_train_glmm_grid <- gridSample(st_coordinates(bg_train_glmm_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") %>% 
  st_join(bg_train_glmm_sf,  # add point atributes 
          join = st_nearest_feature,
          left = TRUE)
nrow(bg_train_glmm_grid)
# [1] 24722

bg_test_glmm_grid <- gridSample(st_coordinates(bg_test_glmm_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") %>% 
  st_join(bg_test_glmm_sf,  # add point atributes 
          join = st_nearest_feature,
          left = TRUE)

nrow(bg_test_glmm_grid)
# [1] 6412

tmap_mode("view")
tm_shape(bg_train_glmm_grid) + #filtered training gps points in color per id 
  tm_dots("id") +
  tm_shape(bg_test_glmm_grid) + #filtered testing gps points in light blue
  tm_dots("lightblue")
  
# write_sf(bg_train_glmm_grid,
#         paste0(dataproc_wd,"/bg_train_bf2018_glmm_3035_AP.gpkg"))
# write_sf(bg_test_glmm_grid,
#         paste0(dataproc_wd,"/bg_test_bf2018_glmm_3035_AP.gpkg"))
```


## Background points for maxent from MCPs.
Before 2018 data
Extract 10x presences number of bg points and divide into training and testing sets

```{r}
# create empty raster based on env raster
raster0 <- raster(env_stack$HFP2009_int)

# create MCP layer with values 0.99 for MCPs
gps_mcp100_bf2018$value <- 0.99
str(gps_mcp100_bf2018)

# give values of 0.01 to cells outside 100% MCPs and 0.99 to 100% MCP cells
raster_mcp_tmp <- fasterize(gps_mcp100_bf2018, raster = raster0, field = "value",  
                            background = 0.01, fun = "min")

plot(raster_mcp_tmp)

# we crop to Germany 
raster_mcp_bg <- mask(raster_mcp_tmp, germany_3035)
plot(raster_mcp_bg)
plot(st_geometry(gps_mcp100_bf2018), add = TRUE)


# extract 10x presences
nrow(gps_bf2018)

bg_mcp_maxent <- xyFromCell(raster_mcp_bg, sample(which(!is.na(values(raster_mcp_bg))), size = nrow(gps_bf2018)*10,
                                         prob=values(raster_mcp_bg)[!is.na(values(raster_mcp_bg))]))


# inspect result by plotting
plot(env_stack$HFP2009_int, legend=FALSE)
plot(st_geometry(gps_mcp100_bf2018), add = TRUE, cex=0.3)
points(bg_mcp_maxent, cex=0.3, col = "red")
# write.csv(bg_mcp_maxent, paste0(dataproc_wd,"/bg_all_bf2018_maxent_mcp_3035_AP.csv"), row.names = FALSE)

head(bg_mcp_maxent)
plot(bg_mcp_maxent)

# divide 4:1 points for train and test
bg_mcp_maxent$fold <- kfold(bg_mcp_maxent, k = 5)
head(bg_mcp_maxent)

bg_train_maxent_mcp <- bg_mcp_maxent[bg_mcp_maxent$fold!=5,] 
bg_test_maxent_mcp <- bg_mcp_maxent[bg_mcp_maxent$fold==5,]

nrow(bg_train_maxent_mcp)
# [1] 25920
nrow(bg_test_maxent_mcp)
# [1] 6480

# make sf objects
bg_train_maxent_mcp_sf <- st_as_sf(bg_train_maxent_mcp, coords = c("x", "y"), crs = 3035) 
bg_test_maxent_mcp_sf <- st_as_sf(bg_test_maxent_mcp, coords = c("x", "y"), crs = 3035) 


# select one observation per grid cell
bg_train_mcp_maxent_grid <- gridSample(st_coordinates(bg_train_maxent_mcp_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") %>% 
  st_join(bg_train_maxent_mcp_sf,  # add point atributes 
          join = st_nearest_feature,
          left = TRUE)
nrow(bg_train_mcp_maxent_grid)
# [1] 25920

bg_test_maxent_mcp_grid <- gridSample(st_coordinates(bg_test_maxent_mcp_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") %>% 
  st_join(bg_test_maxent_mcp_sf,  # add point atributes 
          join = st_nearest_feature,
          left = TRUE)
nrow(bg_test_maxent_mcp_grid)
# [1] 6480


tmap_mode("view")
tm_shape(bg_train_mcp_maxent_grid) + #filtered training gps points in color per id 
  tm_dots("red") +
  tm_shape(bg_test_maxent_mcp_grid) + #filtered testing gps points in light blue
  tm_dots("lightblue")
  
# write_sf(bg_train_mcp_maxent_grid,
#         paste0(dataproc_wd,"/bg_train_bf2018_maxent_mcp_3035_AP.gpkg"))
# write_sf(bg_test_maxent_mcp_grid,
#         paste0(dataproc_wd,"/bg_test_bf2018_maxent_mcp_3035_AP.gpkg"))
```


####################
## territory data ##
####################

These are the centroids of the territories found during wolf monitoring in Germany.

### Load and check territories
```{r}
# load all GPS data for resident adults
terr_sf <- st_read(paste0(dataproc_wd, "/territory_points_3035.gpkg"))
terr_sf
nrow(terr_sf)
# [1] 562

# basic summaries
terr_summary <- terr_sf %>%
  as.data.frame() %>% 
  dplyr::select(-geom) %>% 
  group_by(Mjahr) %>%
  summarise(n = n()) 
terr_summary
# 106 territories in 2018, and 
sum(terr_summary$n) - 106
# 456 territories bf 2018

## Plot the locations color-coded by year
tmap_mode("view")
tm_shape(terr_sf) +
  tm_dots("Mjahr", palette = c("yellow", "orange", "chocolate", "red", "blue"),
          breaks = c(2000, 2005, 2010, 2015, 2018, 2020))
```


## Create bg points for combined model

presences in mcp, bg in territories

The random points here are created using the same bias raster as before, because they use the territories
Only changes the number of random points, now 10x presences in MCPs
We sample random points using a bias raster with 0.99 probability in 10 km buffer around the centroids and 0.01 prob in the rest

```{r}
# Number of presences

pres_points_mcp_maxent_bf2018 <- rbind(pres_train_sf, pres_test_sf)
nrow(pres_points_mcp_maxent_bf2018)
# [1] 3240

# Paralellize the selection of random points (library parallel)
no_cores <- detectCores()/2 - 2 # divide by two detect cores instead of threads
cl <- makeCluster(no_cores)

npoints <- rep(nrow(pres_points_mcp_maxent_bf2018)*10/12, 12)
raster_terr_bg
extent_to_use <- extent(as(germany_3035, "Spatial"))

my_function <- function(npoints){
  dismo::randomPoints(mask = raster_terr_bg,
                      ext = extent_to_use,
                      n = npoints,
                      prob = T)
}

clusterExport(cl, c("raster_terr_bg", "extent_to_use")) # Copy objects to all cores
clusterEvalQ(cl, library("dismo")) # copy libraries to all cores
res.list <- parLapply(cl, npoints, my_function) # Apply parallel functions
stopCluster(cl) # stop parallelization

# Paste all results in the same dataframe
res.df <- do.call("rbind", res.list)
nrow(res.df)
# [1] 32400

res.df2 <- st_as_sf(as.data.frame(res.df), coords = c("x", "y"))
nrow(res.df2)
# 1] 32400

# divide 4:1 points for train and test
res.df2$fold <- kfold(res.df2, k = 5)
head(res.df2)

bg_train_maxent_comb <- res.df2[res.df2$fold!=5,] 
bg_test_maxent_comb <- res.df2[res.df2$fold==5,]

nrow(bg_train_maxent_comb)
# [1] 25920
nrow(bg_test_maxent_comb)
# [1] 6480

# make sf objects
bg_train_maxent_comb_sf <- st_as_sf(bg_train_maxent_comb, coords = c("x", "y"), crs = 3035) 
bg_test_maxent_comb_sf <- st_as_sf(bg_test_maxent_comb, coords = c("x", "y"), crs = 3035) 

# select one observation per grid cell
bg_train_maxent_comb_grid <- gridSample(st_coordinates(bg_train_maxent_comb_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") 

nrow(bg_train_maxent_comb_grid)
# 25843

bg_test_maxent_comb_grid <- gridSample(st_coordinates(bg_test_maxent_comb_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") 
  
nrow(bg_test_maxent_comb_grid)
# 6474

# write_sf(bg_train_maxent_comb_grid,
#          paste0(dataproc_wd,"/bg_train_bf2018_maxent_comb_3035_AP.gpkg"))
# write_sf(bg_test_maxent_comb_grid,
#         paste0(dataproc_wd,"/bg_test_bf2018_maxent_comb_3035_AP.gpkg"))
```


#############################
## evaluation sets >= 2018 ##
#############################

All of the points used as "test set 2"

## GLMM - MCP Individual data

Background points bf 2018 for GLMM from MCPs
Extract 10x presences as bg points test sets 2
```{r}
str(gps_filter_af2018)
nrow(gps_filter_af2018)
# [1] 601

## Calculate 100 MCPs for each individual after 2018
mcp_af2018_dataframe <- gps_filter_af2018 %>%
  as.data.frame() %>%
  dplyr::select(id) %>% 
  droplevels()
mcp_af2018_dataframe <- cbind(mcp_af2018_dataframe, st_coordinates(gps_filter_af2018))
head(mcp_af2018_dataframe)
nrow(mcp_af2018_dataframe)
# [1] 601 # locations before 2018
table(mcp_af2018_dataframe$id) # 5 ids

coordinates(mcp_af2018_dataframe) <- c("X", "Y")
proj4string(mcp_af2018_dataframe) <- CRS( "+init=epsg:3035" )

mcp100_af2018 <- mcp(mcp_af2018_dataframe, percent = 100)
mcp100_af2018
nrow(mcp100_af2018)
# [1] 5
gps_mcp100_af2018 <- st_as_sf(mcp100_af2018, crs = 3035)
plot(st_geometry(gps_mcp100_af2018))
# write_sf(gps_mcp100_af2018, paste0(dataproc_wd, "/gps_MCP100_adults_after2018_3035.gpkg"))

# area of MCPs:
mcp100_af2018_km2 <- st_area(gps_mcp100_af2018)/1000000

mcps_af2018_averages <- mcp100_af2018_km2 %>%
  as.data.frame() %>% 
  summarise(across(everything(), list(mean = mean, sd = sd, 
                                      median = median, 
                                      min = min, max = max))) %>% 
  t()

# add number of points to mcp object
mcp100_af2018_count <- gps_filter_af2018 %>% 
  group_by(id) %>% 
  summarise(n = n()) %>% 
  as.data.frame() %>% 
  dplyr::select(id, n) %>% 
  ungroup() %>% 
  right_join(gps_mcp100_af2018, by = "id") %>% 
  st_sf() %>% 
  as_Spatial() # we need it as spatial object for the next function

# Extract 10 times same amount of background points as presences from MCPs
uni_id <- mcp100_af2018_count$id
bg_points_af2018_tmp <- list()

for(i in 1:length(uni_id)){
  bg_points_af2018_tmp[[i]] <- spsample(mcp100_af2018_count[i,], n = as.numeric(mcp100_af2018_count$n[i])*10, 
                                        type = "random") %>%
    as.data.frame()
  bg_points_af2018_tmp[[i]]$id <- mcp100_af2018_count$id[i] # add territory id
  names(bg_points_af2018_tmp)[i] <- uni_id[i]
}

names(bg_points_af2018_tmp)

bg_points_af2018_tmp2 <- as.data.frame(do.call(rbind, bg_points_af2018_tmp))

# add same color as territories
bg_points_af2018_tmp2 <- bg_points_af2018_tmp2 %>% 
  left_join(as.data.frame(mcp95_all_sf)[,c(1,3)], by = "id") %>% 
  droplevels() # remove id with 0 observations

# make sf objects
bg_test_af2018_sf <- st_as_sf(bg_points_af2018_tmp2, coords = c("x", "y"), crs = 3035) 

# select one observation per grid cell
# extract coordinates of 1 observation per grid cell
bg_test_af2018_glmm_grid <- gridSample(st_coordinates(bg_test_af2018_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") %>% 
  st_join(bg_test_af2018_sf,  # add point atributes 
          join = st_nearest_feature,
          left = TRUE)
nrow(bg_test_af2018_glmm_grid)
# [1] 5805

tmap_mode("plot")
tm_shape(as_Spatial(mcp95_all_sf)) +
  tm_borders(col = mcp95_all_sf$color) +
tm_shape(as_Spatial(bg_test_af2018_glmm_grid)) + #filtered test after 2018 points  
  tm_dots(col = bg_test_af2018_glmm_grid$color) 

# get colors with 0.2 alpha manually (Change last FF for 33)
my_colors <- paste0(str_sub(mcp95_all_sf$color, 1, 7), "33")

plot(st_geometry(mcp95_all_sf), col = my_colors)
plot(st_geometry(bg_test_af2018_glmm_grid), col = bg_test_af2018_glmm_grid$color, 
     pch = 1, cex = 0.1, add = TRUE)

# write_sf(bg_test_af2018_glmm_grid,
#         paste0(dataproc_wd,"/bg_test2_af2018_glmm_3035_AP.gpkg"))
```


## Maxent / Ensemble - MCP data
```{r}
# create a raster to assign probability values in the cells, using environmental data as template
raster0 <- raster(env_stack$HFP2009_int)

# create MCP layer with values 0.99 for MCPs
gps_mcp100_af2018$value <- 0.99
str(gps_mcp100_af2018)

# give values of 0.01 to cells outside 100% MCPs and 0.99 to 100% MCP cells
raster_mcp_af2018_tmp <- fasterize(gps_mcp100_af2018, raster = raster0, field = "value",  
                            background = 0.01, fun = "max")

plot(raster_mcp_af2018_tmp)

# we crop to Germany 
raster_mcp_af2018_bg <- mask(raster_mcp_af2018_tmp, germany_3035)
plot(raster_mcp_af2018_bg)
plot(st_geometry(gps_mcp100_af2018), add = TRUE)

# extract 10x presences
nrow(gps_filter_af2018) #[1] 601

bg_mcp_af2018_maxent <- xyFromCell(raster_mcp_af2018_bg, sample(which(!is.na(values(raster_mcp_af2018_bg))), 
                                                                size = nrow(gps_filter_af2018)*10,
                                                                prob=values(raster_mcp_af2018_bg)[!is.na(values(raster_mcp_af2018_bg))]))

# inspect result by plotting
plot(env_stack$HFP2009_int, legend=FALSE)
plot(st_geometry(gps_mcp100_af2018), add = TRUE, cex=0.3)
points(bg_mcp_af2018_maxent, cex=0.3, col = "red")

nrow(bg_mcp_af2018_maxent) # [1] 6010

head(bg_mcp_af2018_maxent)
plot(bg_mcp_af2018_maxent)

# make sf objects
bg_test2_af2018_maxent_mcp_sf <- bg_mcp_af2018_maxent %>% 
  as.data.frame() %>% 
  st_as_sf(coords = c("x", "y"), crs = 3035)

# select one observation per grid cell
bg_test2_af2018_mcp_maxent_grid <- gridSample(st_coordinates(bg_test2_af2018_maxent_mcp_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") 
nrow(bg_test2_af2018_mcp_maxent_grid)
# [1] 6010

tm_shape(mcp100_af2018) + #filtered training gps points in color per id 
  tm_dots("red") +
  tm_shape(bg_test2_af2018_mcp_maxent_grid) + #filtered testing gps points in light blue
  tm_dots("lightblue")
  
# write_sf(bg_test2_af2018_mcp_maxent_grid,
#         paste0(dataproc_wd,"/bg_test2_af2018_maxent_mcp_3035_AP.gpkg"))
```

## Maxent / ensemble - Combined data

presences in mcp, bg in territories

The random points here are created using the same bias raster as before, because they use the territories
Only changes the number of random points, now 10x presences in MCPs
We sample random points using a bias raster with 0.99 probability in 10 km buffer around the centroids and 0.01 prob in the rest

```{r}
# Load presences
gps_filter_af2018 <- st_read(paste0(dataproc_wd, "/gps_points_eval_af2018_3035.gpkg"))
raster_terr_bg_af2018 <- raster(paste0(workDir, "/output/geo-proc/raster_bias_terr_af2018_100m_3035.tif"))

nrow(gps_filter_af2018)
# [1] 601

# Paralellize the selection of random points (library parallel)
no_cores <- detectCores()/2 - 2 # divide by two detect cores instead of threads
cl <- makeCluster(no_cores)

npoints <- rep(nrow(gps_filter_af2018), 10)
raster_terr_bg_af2018
extent_to_use <- extent(as(germany_3035, "Spatial"))

my_function <- function(npoints){
  dismo::randomPoints(mask = raster_terr_bg_af2018,
                      ext = extent_to_use,
                      n = npoints,
                      prob = T)
}

clusterExport(cl, c("raster_terr_bg_af2018", "extent_to_use")) # Copy objects to all cores
clusterEvalQ(cl, library("dismo")) # copy libraries to all cores
res.list <- parLapply(cl, npoints, my_function) # Apply parallel functions
stopCluster(cl) # stop parallelization

# Paste all results in the same dataframe
res.df <- do.call("rbind", res.list)
nrow(res.df)
# [1] 6010

res.df2 <- st_as_sf(as.data.frame(res.df), coords = c("x", "y"))
nrow(res.df2)
# 1] 6010

# make sf objects
bg_test2_af2018_maxent_comb_sf <- st_as_sf(res.df2, coords = c("x", "y"), crs = 3035) 

# select one observation per grid cell
bg_test2_af2018_maxent_comb_grid <- gridSample(st_coordinates(bg_test2_af2018_maxent_comb_sf), # select 1 point per grid
                           copernicus_lc_3035, 
                           n = 1) %>%
  as.data.frame() %>% 
  st_as_sf(coords = c("X","Y"), # make sf object again
           crs = 3035, 
           sf_column_name = "geometry") 

nrow(bg_test2_af2018_maxent_comb_grid)
# 6009

# inspect result by plotting
# plot(env_stack$HFP2009_int, legend=FALSE)
plot(germany_3035, add = TRUE)
plot(st_geometry(bg_test2_af2018_maxent_comb_grid), add = TRUE, cex=0.3, col = "red")

# write_sf(bg_test2_af2018_maxent_comb_grid,
#          paste0(dataproc_wd,"/bg_test2_af2018_maxent_comb_3035_AP.gpkg"))
```



