---
title: "env_exploration"
author: "Aimara Planillo"
date: "2/8/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploring environmental layers

```{r prepare workspace}
## packages
source("./Rscripts/source_file_packages.R")

## work directories
workDir <- getwd()

dataproc_wd <- paste0(workDir, "/output/data_proc")
geoproc_wd <- paste0(workDir, "/output/geo_proc")
results_wd <- paste0(workDir, "/output/results")
```


### Load environmental layers
```{r}
## Border of Germany
germany_3035 <- read_sf(paste0(geoproc_wd, "/germany_3035.gpkg"), crs = 3035) 

## env stack
env_stack <- stack(paste0(geoproc_wd, "/stack_all_env_vars_20181220.tif"))
# add names to the stack
names(env_stack) <- c("g100_clc12_V18_5", "popu01clcv5", "mos2_dist_settlments", "mos2_dist_streets", "HFP2009_int")

# Correct land use classes to seven classes
env_stack$g100_clc12_V18_5[env_stack$g100_clc12_V18_5 == 103] <- 102

unique(env_stack$g100_clc12_V18_5)

plot(env_stack$mos2_dist_settlments)

plot(st_geometry(germany_3035), add = TRUE)

# mask the environmental layers to Germany and 10km buffer around
env_germany <- mask(env_stack, germany_3035 %>% st_buffer(10000))

# rename and reorganize to follow the order of the paper
names(env_germany) <- c("CLC2012", "Pop_den", "Dist_settl", "Dist_roads", "HFP") 


plot(env_germany)
plot(env_germany$Dist_settl)
plot(st_geometry(germany_3035), add = TRUE)
# unique(env_germany$CLC2012)

plot(env_germany$Dist_roads)
plot(st_geometry(germany_3035), add = TRUE)


# rename and reorganize to follow the order of the paper
env_germany <- stack(env_germany[[1]],env_germany[[2]],env_germany[[5]],
                    env_germany[[3]], env_germany[[4]])

# writeRaster(env_germany, paste0(geoproc_wd, "/stack_env_rasters_germany_3035_AP.tif"))
```


### Correlation among the layers
```{r}
# pairwise correlation of the rasters
rasters_cor <- layerStats(env_germany, 'pearson', na.rm=T)
corr_matrix <- rasters_cor$'pearson correlation coefficient'
corr_matrix

corr_matrix[corr_matrix > 0.5 | corr_matrix < -0.5] 

# save data
write.csv(corr_matrix, paste0(results_wd, "/env_corr_matrix_pearson.csv"), row.names = TRUE)
corr_matrix <- read.csv(paste0(results_wd, "/env_corr_matrix_pearson.csv"), row.names = 1) %>%  as.matrix()
```



# Extract env variables for presences

We need this explicitly for the GLMM models, 
but we do it for all to get the values for the summary tables

## Presence points 
## Env cov for Presences bf 2018
```{r}
# load GPS telemetry presences
pres_train_gps_sf <- st_read(paste0(dataproc_wd,"/gps_points_train_bf2018_3035.gpkg")) %>% 
  st_transform(crs = 3035) %>% 
  mutate(type = "train")
pres_test_gps_sf <- st_read(paste0(dataproc_wd,"/gps_points_test_bf2018_3035.gpkg")) %>% 
  st_transform(crs = 3035) %>% 
  mutate(type = "test")

pres_gps <- rbind(pres_train_gps_sf, pres_test_gps_sf)
pres_gps
plot(st_geometry(pres_gps))
nrow(pres_gps) #[1] 3240

# extract env cov for the points
pres_gps_envcov <- raster::extract(env_germany, pres_gps, sp = TRUE)
summary(pres_gps_envcov)
nrow(pres_gps_envcov) #[1] 3240

pres_gps_envcov_sf <- st_as_sf(pres_gps_envcov)
# write_sf(pres_gps_envcov_sf, paste0(dataproc_wd, "/envcov_pres_telemetry.gpkg"))

## Get the summaries
pres_gps_envcov_sf$CLC2012[which.max(pres_gps_envcov_sf$CLC2012)]

(summary_pres_gps_envcov <- pres_gps_envcov_sf %>%
  as.data.frame() %>% 
  group_by(type) %>%
  summarise(npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
  mutate(data_source = "telemetry_filtered_presences")
)
```

## Env cov for Presences af 2018
```{r}
# load GPS telemetry presences
gps_mcp_test2_af2018 <- st_read(paste0(dataproc_wd, "/gps_points_eval_af2018_3035.gpkg")) %>% 
  st_transform(crs = 3035)
  
nrow(gps_mcp_test2_af2018) #[1] 601

# extract env cov for the points
pres_gps_af2018_envcov <- raster::extract(env_germany, gps_mcp_test2_af2018, sp = TRUE)
summary(pres_gps_af2018_envcov)
nrow(pres_gps_af2018_envcov) #[1] 601

pres_gps_af2018_envcov_sf <- st_as_sf(pres_gps_af2018_envcov)
# write_sf(pres_gps_af2018_envcov_sf, paste0(dataproc_wd, "/envcov_pres_telemetry_af2018.gpkg"))

## Get the summaries
(summary_pres_gps_af2018_envcov <- pres_gps_af2018_envcov_sf %>%
  as.data.frame() %>% 
  # group_by(type) %>%
  summarise(type = "test2", 
            npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
  mutate(data_source = "telemetry_filtered_presences_af2018")
)
```


## Background points 

### Env cov for background mcp for glmm bf 2018

```{r}
# Load Background points MCP GLMM
bg_train_glmm_sf <- st_read(paste0(dataproc_wd,"/bg_train_bf2018_glmm_3035_AP.gpkg")) %>% 
  st_transform(crs = 3035) %>% 
  mutate(type = "train")
plot(germany_3035)
plot(st_geometry(bg_train_glmm_sf), add = TRUE)
bg_test_glmm_sf <- st_read(paste0(dataproc_wd,"/bg_test_bf2018_glmm_3035_AP.gpkg")) %>% 
  st_transform(crs = 3035) %>% 
  mutate(type = "test")

bg_mcp_glmm <- rbind(bg_train_glmm_sf, bg_test_glmm_sf)
bg_mcp_glmm
plot(st_geometry(bg_mcp_glmm))
nrow(bg_mcp_glmm) #[1] 31134

bg_mcp_glmm_envcov <- raster::extract(env_germany, bg_mcp_glmm, sp = TRUE)
summary(bg_mcp_glmm_envcov)
nrow(bg_mcp_glmm_envcov) #[1] 31134

bg_mcp_glmm_envcov_sf <- st_as_sf(bg_mcp_glmm_envcov)
# write_sf(bg_mcp_glmm_envcov_sf, paste0(dataproc_wd, "/envcov_bg_mcp_glmm.gpkg"))

## Get the summaries
(summary_bg_mcp_glmm_envcov <- bg_mcp_glmm_envcov_sf %>%
  as.data.frame() %>% 
  group_by(type) %>%
  summarise(npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
  mutate(data_source = "bg_mcp_glmm")
)
```

### Env cov for background mcp for telemetry maxent / ensemble bf 2018

```{r}
# Load Background points MCP maxent / ensemble
bg_train_maxent_mcp_sf <- st_read(paste0(dataproc_wd,"/bg_train_bf2018_maxent_mcp_3035_AP.gpkg")) %>% 
   st_transform(crs = 3035) %>% 
  mutate(type = "train")
bg_test_maxent_mcp_sf <- st_read(paste0(dataproc_wd,"/bg_test_bf2018_maxent_mcp_3035_AP.gpkg")) %>% 
   st_transform(crs = 3035) %>% 
  mutate(type = "test")

bg_mcp_maxent <- rbind(bg_train_maxent_mcp_sf, bg_test_maxent_mcp_sf)
bg_mcp_maxent
plot(st_geometry(bg_mcp_maxent))
nrow(bg_mcp_maxent) #[1] 32400

bg_mcp_maxent_envcov <- raster::extract(env_germany, bg_mcp_maxent, sp = TRUE)
summary(bg_mcp_maxent_envcov)
nrow(bg_mcp_maxent_envcov) #[1] 32400

bg_mcp_maxent_envcov_sf <- st_as_sf(bg_mcp_maxent_envcov)
# write_sf(bg_mcp_maxent_envcov_sf, paste0(dataproc_wd, "/envcov_bg_mcp_maxent.gpkg"))


## Get the summaries
(summary_bg_mcp_maxent_envcov <- bg_mcp_maxent_envcov_sf %>%
  as.data.frame() %>% 
  group_by(type) %>%
  summarise(npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
    mutate(data_source = "bg_mcp_maxent")
)
```

### Env cov for background mcp for combined maxent / ensemble bf 2018

```{r}
# Load Background points combined maxent
bg_train_maxent_comb_sf <- st_read(paste0(dataproc_wd,"/bg_train_bf2018_maxent_comb_3035_AP.gpkg")) %>% 
   st_transform(crs = 3035) %>% 
  mutate(type = "train")
bg_test_maxent_comb_sf <- st_read(paste0(dataproc_wd,"/bg_test_bf2018_maxent_comb_3035_AP.gpkg")) %>% 
   st_transform(crs = 3035) %>% 
  mutate(type = "test")

bg_comb_maxent <- rbind(bg_train_maxent_comb_sf, bg_test_maxent_comb_sf)
bg_comb_maxent
plot(st_geometry(bg_comb_maxent))
nrow(bg_comb_maxent) #[1] 32317

bg_comb_maxent_envcov <- raster::extract(env_germany, bg_comb_maxent, sp = TRUE)
summary(bg_comb_maxent_envcov)
nrow(bg_comb_maxent_envcov) #[1] 32317

bg_comb_maxent_envcov_sf <- st_as_sf(bg_comb_maxent_envcov)
# write_sf(bg_comb_maxent_envcov_sf, paste0(dataproc_wd, "/envcov_bg_combined_maxent.gpkg"))

## Get the summaries
(summary_bg_comb_maxent_envcov <- bg_comb_maxent_envcov_sf %>%
  as.data.frame() %>% 
  group_by(type) %>%
  summarise(npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
    mutate(data_source = "bg_comb_maxent")
)
```

### Env cov for background points after 2018

```{r}
# Load Background points MCP GLMM
bg_test_af2018_glmm_sf <- st_read(paste0(dataproc_wd,"/bg_test2_af2018_glmm_3035_AP.gpkg"))
nrow(bg_test_af2018_glmm_sf) #[1] 5805

bg_mcp_glmm_test2_envcov <- raster::extract(env_germany, bg_test_af2018_glmm_sf, sp = TRUE)
summary(bg_mcp_glmm_test2_envcov)
nrow(bg_mcp_glmm_test2_envcov) #[1] 5805

bg_mcp_af2018_glmm_envcov_sf <- st_as_sf(bg_mcp_glmm_test2_envcov)
# write_sf(bg_mcp_glmm_envcov_sf, paste0(dataproc_wd, "/envcov_bg_mcp_glmm_af2018.gpkg"))

## Get the summaries
(summary_bg_mcp_glmm_af2018_envcov <- bg_mcp_af2018_glmm_envcov_sf %>%
  as.data.frame() %>% 
  # group_by(type) %>%
  summarise(type = "test2", 
            npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
  mutate(data_source = "bg_mcp_glmm_af2018")
)


# Load Background points MCP maxent / ensemble
bg_test_af2018_maxent_mcp_sf <- st_read(paste0(dataproc_wd,"/bg_test2_af2018_maxent_mcp_3035_AP.gpkg")) %>% 
   st_transform(crs = 3035) 
nrow(bg_test_af2018_maxent_mcp_sf) #[1] 6010

bg_af2018_mcp_maxent_envcov <- raster::extract(env_germany, bg_test_af2018_maxent_mcp_sf, sp = TRUE)
summary(bg_af2018_mcp_maxent_envcov)
nrow(bg_af2018_mcp_maxent_envcov) #[1] 6010

bg_af2018_mcp_maxent_envcov_sf <- st_as_sf(bg_af2018_mcp_maxent_envcov)
# write_sf(bg_af2018_mcp_maxent_envcov_sf, paste0(dataproc_wd, "/envcov_bg_mcp_maxent_af2018.gpkg"))

## Get the summaries
(summary_bg_mcp_maxent_af2018_envcov <- bg_af2018_mcp_maxent_envcov_sf %>%
  as.data.frame() %>% 
  # group_by(type) %>%
  summarise(type = "test2", 
            npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
    mutate(data_source = "bg_mcp_maxent_af2018")
)


# Load Background points af2018 combined maxent
bg_test2_af2018_maxent_comb_sf <- st_read(paste0(dataproc_wd,"/bg_test2_af2018_maxent_comb_3035_AP.gpkg")) %>% 
   st_transform(crs = 3035) 
nrow(bg_test2_af2018_maxent_comb_sf) #[1] 6009

bg_af2018_comb_maxent_envcov <- raster::extract(env_germany, bg_test2_af2018_maxent_comb_sf, sp = TRUE)
summary(bg_af2018_comb_maxent_envcov)
nrow(bg_af2018_comb_maxent_envcov) #[1] 6009

bg_af2018_comb_maxent_envcov_sf <- st_as_sf(bg_af2018_comb_maxent_envcov)
# write_sf(bg_af2018_comb_maxent_envcov_sf, paste0(dataproc_wd, "/envcov_bg_comb_maxent_af2018.gpkg"))

## Get the summaries
(summary_bg_comb_maxent_af2018_envcov <- bg_af2018_comb_maxent_envcov_sf %>%
  as.data.frame() %>% 
  # group_by(type) %>%
  summarise(type = "test2", 
            npoints = n(), 
            land_use = getmode(CLC2012),
            across(Pop_den:Dist_roads, 
                   list(min = min, max = max, mean = mean, 
                        median = median, sd = sd), na.rm= TRUE)) %>% 
    mutate(data_source = "bg_comb_maxent_af2018")
)
```



## Summary of all points

```{r}
summary_allpoints <- rbind(summary_pres_gps_envcov, summary_pres_gps_af2018_envcov, 
                           summary_bg_mcp_glmm_envcov, summary_bg_mcp_maxent_envcov,
                           summary_bg_comb_maxent_envcov, 
                           summary_bg_mcp_glmm_af2018_envcov, summary_bg_mcp_maxent_af2018_envcov,
                           summary_bg_comb_maxent_af2018_envcov)

# write.csv(summary_allpoints, paste0(dataproc_wd, "/Table_envcov_extract_summary.csv"),
#          row.names = FALSE)
```

```{r get summaries per variable}
summary_allpoints[,grep("land_use", colnames(summary_allpoints))]
summary_allpoints[,grep("Pop", colnames(summary_allpoints))]
summary_allpoints[,grep("HFP", colnames(summary_allpoints))]
summary_allpoints[,grep("sett", colnames(summary_allpoints))]
summary_allpoints[,grep("roads", colnames(summary_allpoints))]
```


## Plot env values to inspect them
```{r}
# check values 
summary(env_germany)

# fast plotting: library quickplot
Plot(env_germany, new = TRUE)

# Highlight NA values in black
Plot(env_germany)

clearPlot()

# give color palette for each map
mapColours <- list(
  CLC2012 = brewer.pal(8, "Spectral"), 
  Pop_den = brewer.pal(8, "Spectral"), 
  HFP = brewer.pal(8, "Spectral"), 
  Dist_settl = brewer.pal(8, "Spectral"), 
  Dist_roads = brewer.pal(8, "Spectral"))
setColors(env_germany, n = 8) <- mapColours

Plot(env_germany, new = TRUE, na.color = "black")
```


