---
title: "ensemble model"
author: "Aimara Planillo"
date: "3/11/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### WORKSPACE

```{r}
# libraries and functions
source("./R/source_file.R")

workDir <- getwd()
dataproc_wd <- paste0(workDir, "/output/data_proc")
geoproc_wd <- paste0(workDir, "/output/geo_proc")
results_wd <- paste0(workDir, "/output/ensemble_global")
```

**Note**: In this case we use an old code suitable for biomod2 v 3.4

# Ensemble model for Wolf Habitat Suitability in Germany

Based on the results of the model comparison and performance on the test data and the 
extra evaluation set, we are going to use the telemetry data as presences and the 
territory buffers from the monitoring as background, with data from all years together
to obtain the most robust model to predict habitat suitability for the establishment 
of wolf territories in Germany.

The ensemble model is the method selected because it combines the strengths of multiple
modelling techniques, given a consensus model built from all individual models, weighting 
them by their accuracy, in this case AUC.

We follow the workflow that was already proposed in Biomod2


### Load data
```{r}
# environmental rasters
env_germany <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_3035_AP.tif"))
names(env_germany) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads") 

### Wolf telemetry data (all years) ###

# Load observations with env data bf 2018
pres_gps_bf2018_envcov_sf <- st_read(paste0(dataproc_wd, "/envcov_pres_telemetry.gpkg")) %>% 
  st_transform(crs = 3035)  
table(pres_gps_envcov_sf$type) 

# Load observation af 2018
pres_gps_af2018_envcov_sf <- st_read(paste0(dataproc_wd,"/envcov_pres_telemetry_af2018.gpkg")) %>% 
  st_transform(crs = 3035)
nrow(pres_gps_af2018_envcov_sf) 

# Put all presences in one data frame
colnames(pres_gps_bf2018_envcov_sf)
colnames(pres_gps_af2018_envcov_sf)

wolf_pres <- pres_gps_bf2018_envcov_sf %>% 
  dplyr::select(-c(type, fold)) %>% 
  rbind(pres_gps_af2018_envcov_sf)

wolf_pres 

### Background data ###
bg_comb_bf2018 <- st_read(paste0(dataproc_wd,"/envcov_bg_combined_maxent.gpkg"))
bg_comb2_af2018 <- st_read(paste0(dataproc_wd,"/envcov_bg_comb_maxent_af2018.gpkg")) %>% 
   st_transform(crs = 3035) 

nrow(bg_comb_bf2018) 
nrow(bg_comb2_af2018) 

# Put all background in one data frame
colnames(bg_comb_bf2018)
colnames(bg_comb2_af2018)

wolf_bg <- bg_comb_bf2018 %>% 
  dplyr::select(-c(type)) %>% 
  rbind(bg_comb2_af2018) 

wolf_bg 

### Divide into train and test data (4:1 - 80%-20%)
# Presences
wolf_pres$fold <- kfold(wolf_pres, k = 5)
wolf_pres <- wolf_pres %>% 
  mutate(type = ifelse(fold == 5, "test", "train"), 
         data = "presences")
  
wolf_pres_train <- wolf_pres %>% 
  filter(fold != 5)
wolf_pres_test <- wolf_pres %>% 
  filter(fold == 5)
nrow(wolf_pres_train) 
nrow(wolf_pres_test) 

## Select same number of points as number of presences
wolf_bg$id <- seq(1:nrow(wolf_bg))
bg_sample <- sample(wolf_bg$id, nrow(wolf_pres))

wolf_bg <- wolf_bg[wolf_bg$id %in% bg_sample,]

# Background
wolf_bg$fold <- kfold(wolf_bg, k = 5)
wolf_bg <- wolf_bg %>% 
  dplyr::select(-id) %>% 
  mutate(type = ifelse(fold == 5, "test", "train"), 
         data = "background")

wolf_bg_train <- wolf_bg %>% 
  filter(fold != 5)
wolf_bg_test <- wolf_bg %>% 
  filter(fold == 5)
nrow(wolf_bg_train) 
nrow(wolf_bg_test) 

### Put all togheter and safe
head(wolf_pres)
head(wolf_bg)

all_wolf_points <- wolf_pres %>% 
  dplyr::select(CLC2012:data) %>% 
  rbind(wolf_bg)

str(all_wolf_points)
table(all_wolf_points$data, all_wolf_points$type)

# st_write(all_wolf_points, paste0(dataproc_wd, "/all_wolf_points_samebgpres_ensemble_3035.gpkg"))
```


## Run models

```{r}
## Formating data
# Transform points into presence (1) - absence (0) and divide into train and test sets
wolf_dat_train <- all_wolf_points %>% 
  filter(fold != 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()
wolf_dat_train
table(wolf_dat_train$pres_bg)

nrow(wolf_dat_train) 

wolf_dat_test <- all_wolf_points %>% 
  filter(fold == 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()

wolf_dat_test
table(wolf_dat_test$pres_bg)

nrow(wolf_dat_test) 

## Extract environmental covariates
head(all_wolf_points)

my_covariates_train <- all_wolf_points %>%
  filter(fold != 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(my_covariates_train)
str(my_covariates_train) 

my_covariates_test <- all_wolf_points %>%
  filter(fold == 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(my_covariates_test)
str(my_covariates_test) 

## Specify biomod objects
wolf_mod_dat <- BIOMOD_FormatingData(resp.var = wolf_dat_train,
                                    expl.var = my_covariates_train,
                                    resp.name = "Wolf_2")
wolf_mod_dat
                                    
                                    
## Define models
# For maxent, we need to add some options to the models. 
# Important, the directory where to find java/maxent. 
myBiomodOptions <- BIOMOD_ModelingOptions(
  MAXENT.Phillips = list( path_to_maxent.jar = "C:/TEMP/maxent.jar",
                          maximumiterations = 5000,
                          responsecurves=TRUE,
                          jacknife=TRUE,
                          replicates=3,
                          visible = FALSE,
                          directory="/Wolf_2/",
                          linear = TRUE, 
                          quadratic = TRUE,
                          product = TRUE,
                          threshold = FALSE,
                          hinge = TRUE,
                          lq2lqptthreshold = 80,
                          l2lqthreshold = 10,
                          hingethreshold = 15,
                          beta_threshold = -1,
                          beta_categorical = -1,
                          beta_lqp = -1,
                          beta_hinge = -1,
                          defaultprevalence = 0.5))

wolf_mod_out <- BIOMOD_Modeling(data = wolf_mod_dat,
                            models =c('GLM','GAM','GBM', 'MARS', 'ANN','RF','MAXENT.Phillips'),
                            SaveObj = TRUE,
                           models.options = myBiomodOptions,
                           NbRunEval = 10,
                           DataSplit = 80, # common practice to validate: 80 train, 20 test
                           models.eval.meth = c('KAPPA','TSS','ROC'),
                           VarImport = 10)

wolf_mod_out
str(wolf_mod_out)
wolf_mod_out@models.computed
# saveRDS(wolf_mod_out, paste0(results_wd, "/all_models_wolf_samebgpres_train80.rds"))

# get all models evaluation
wolf_ModelEval <- get_evaluations(wolf_mod_out)
wolf_ModelEval
# print the dimnames of this object
dimnames(wolf_ModelEval)

# get the AUC scores 
(AUC_table <- wolf_ModelEval["ROC","Testing.data",,,])
rowSums(AUC_table[,1:10])/10

# get the TSS scores 
(TSS_table <- wolf_ModelEval["TSS","Testing.data",,,])
rowSums(TSS_table[,1:10])/10

# write.csv(AUC_table, paste0(results_wd, "/all_models_samebgpres_AUCTable.csv"))
# write.csv(TSS_table, paste0(results_wd, "/all_models_samebgpres_TSSTable.csv"))
```

## get variable importance
For the machine learning algorithms ANN and GBM, the AIC approach is not applicable because there is no model log-likelihood information available. Therefore, we have included the variable importance function that is implemented in the biomod2 package. This function uses a machine-learning approach once the models are trained to randomize one of the variables in each permutation and calculate a correlation score between the standard prediction and the new prediction. This score is considered to give an estimation of the variable importance in the model. The higher the value, the more importance the predictor variable has on the model. A value of 0 assumes no influence of that predictor. Note that this method does not account for interactions between variables and should be considered more as an information tool for each model independently.
```{r}
# print variable importances
wolf_VarImpor <- get_variables_importance(wolf_mod_out)
wolf_VarImpor
# Variable importance of the full model
VarImpor_table <- wolf_VarImpor[,,"Full","AllData"]
VarImpor_table
# write.csv(VarImpor_table, paste0(results_wd, "/all_models_samebgpres_VariableImportance_FullModel.csv"))


## Average variable importance in the 10 model replicates
# extract the matrices to a list
wolf_VarImpor_list <- list()
for (i in 1:10){
  wolf_VarImpor_list[[i]] <- as.data.frame(wolf_VarImpor[,,i,])
}

for (i in 1:length(wolf_VarImpor_list)){
  wolf_VarImpor_list[[i]]$RUN <- paste0("Run_", i)
}

wolf_VarImport_allmodels <- do.call(rbind, wolf_VarImpor_list)
# write.csv(wolf_VarImport_allmodels, paste0(results_wd, "/all_models_samebgpres_VariableImportance.csv"))


##Varible contributions for Maxent, not possible this method for other models
forSetup <- read.csv(file = paste("./Wolf.2/models/1618066184/Wolf.2_AllData_Full_MAXENT.Phillips_outputs/maxentResults.csv", 
                                  sep = ""), header = T)#Choose the appropriate model folder with the seed of the analysis you want
variableContributions <- matrix(data = NA, nrow = length(forSetup[, grep('.contribution', names(forSetup))]), ncol = 10);
rownames(variableContributions) <- names(forSetup[, grep('.contribution', names(forSetup))])
colnames(variableContributions) <- c("Run1", "Run2", "Run3", "Run4", "Run5", "Run6", "Run7", "Run8", "Run9", "Run10")
variablePermutationImportance <- matrix(data = NA, nrow = length(forSetup[, grep('.permutation.importance', names(forSetup))]), ncol = 10);
colnames(variablePermutationImportance) <- c("Run1", "Run2", "Run3", "Run4", "Run5", "Run6", "Run7", "Run8", "Run9", "Run10")
count <- 1;
while (count <= 10){
  temporary <- read.csv(file = paste("./Wolf/models/1617188603/Wolf_AllData_RUN", count, "_MAXENT.Phillips_outputs/maxentResults.csv", sep = ""), header = T);
  variableContributions[,count] <- unlist(unname(temporary[, grep('.contribution', names(temporary))]))
  variablePermutationImportance[,count] <- unlist(unname(temporary[, grep('.permutation.importance', names(temporary))]))
  count <- count + 1;
}
# write.csv(variableContributions, paste0(results_wd, "/Maxent_samebgpres_VariableContributions.csv"))
# write.csv(variablePermutationImportance, paste0(results_wd, "/Maxent_VariablePermutationImportance.csv"))


# response curves
## Variable response curves
response.plot2(models = BIOMOD_LoadModels(wolf_mod_out, models='MAXENT.Phillips'),
               Data = get_formal_data(wolf_mod_out,'expl.var'),
               show.variables= get_formal_data(wolf_mod_out,'expl.var.names'),
               do.bivariate = FALSE,
               fixed.var.metric = 'mean',
               col = brewer.pal(10, "Spectral"),
               legend = TRUE,
               data_species = get_formal_data(wolf_mod_out,'resp.var'))
```


## Make ensemble
```{r}
### Building ensemble-models
# All models with AUC > 0.7
wolf_ensemblemod <- BIOMOD_EnsembleModeling(
  modeling.output = wolf_mod_out,
  chosen.models = 'all',
  em.by='all', 
  eval.metric = c('ROC'),
  eval.metric.quality.threshold = c(0.7), # model above 0.7 AUC values
  prob.mean = T,
  prob.cv = T,
  prob.ci = T,
  prob.ci.alpha = 0.05,
  prob.median = T,
  committee.averaging = T,
  prob.mean.weight = T,
  prob.mean.weight.decay = 'proportional')

wolf_ensemblemod
# saveRDS(wolf_ensemblemod, paste0(results_wd, "/ensemble_wolf_samebgpres.rds"))
wolf_ensemblemod <- readRDS(paste0(results_wd, "/ensemble_results/ensemble_wolf_samebgpres.rds"))
# get evaluation scores
scores_all <- get_evaluations(wolf_ensemblemod)
threshold <- scores_all$Wolf.2_EMwmeanByROC_mergedAlgo_mergedRun_mergedData["TSS", "Cutoff"]
threshold
```


## Project models
```{r}
### Project individual models
wolf_BiomodProj <- BIOMOD_Projection(
  modeling.output = wolf_mod_out,
  new.env = env_germany,
  proj.name = 'ind_models',
  selected.models = 'all',
  compress = 'gzip',
  build.clamping.mask = F,
  output.format = '.grd', 
  do.stack=T)

wolf_BiomodProj
# saveRDS(wolf_BiomodProj, paste0(results_wd, "/all_models_samebgpres_biomod_projection.rds"))

names(wolf_BiomodProj)


# make some plots sub-selected by str.grep argument
plot(wolf_BiomodProj, str.grep ='RF')

# to make custom plots, get the projected map
wolf_indmodelProj <- get_predictions(wolf_BiomodProj)
wolf_indmodelProj
wolf_Result <- calc(wolf_indmodelProj,fun = mean); #Choose whatever descriptive statistic you'd like
plot(wolf_Result)
# writeRaster(wolf_Result, filename = paste0(results_wd, "/all_models_samebgpres_mean_projection_3035.tif"),
#             overwrite = T);

## Project ensemble-model
wolf_BiomodEF <- BIOMOD_EnsembleForecasting(EM.output = wolf_ensemblemod,
                            projection.output = wolf_BiomodProj,
                            selected.models = 'all',
                            compress = 'gzip')
wolf_BiomodEF
# saveRDS(wolf_BiomodEF, paste0(results_wd, "/ensemble_samebgpres_forcasting.rds"))

wolf_projEnsemble <- get_predictions(wolf_BiomodEF)
names(wolf_projEnsemble)
wolf_EnsembleResult <- wolf_projEnsemble[[1]] #This is the mean model ensemble across all models
wolf_EnsembleResult_wm <- wolf_projEnsemble$Wolf.2_EMwmeanByROC_mergedAlgo_mergedRun_mergedData # weighted mean model based on AUC
plot(wolf_EnsembleResult)
plot(wolf_EnsembleResult_wm)

# writeRaster(wolf_EnsembleResult, filename = paste0(results_wd, 
#                                                    "/ensemble_samebgpres_mean_projection_3035.tif"),
#             overwrite = T)
# writeRaster(wolf_EnsembleResult_wm,
#             filename = paste0(results_wd, "/ensemble_samebgpres_weightedmean_projection_3035.tif"),
#             overwrite = T)
```


## Plot individual models
```{r}
# plot fancy map for MAXENT
wolf_Maxent_pred <- wolf_indmodelProj$Wolf.2_AllData_Full_MAXENT.Phillips
# writeRaster(wolf_Maxent_pred, filename = paste0(results_wd,
#                                                 "/all_models_samebgpres_Maxent_projection_3035.tif"),
#             overwrite = T)


# convert to a df for plotting in two steps,
# First, to a SpatialPointsDataFrame
wolf_maxent_pts <- rasterToPoints(wolf_Maxent_pred, spatial = TRUE)
# Then to a 'conventional' dataframe
wolf_maxent_df  <- data.frame(wolf_maxent_pts)
wolf_maxent_df$maxent <- wolf_maxent_df$Wolf.2_AllData_Full_MAXENT.Phillips/1000

wolf_maxent_map <- ggplot(data = wolf_maxent_df, aes(y=y, x=x)) +
  geom_raster(aes(fill=maxent)) +
  theme_bw() +
  coord_equal() +
  scale_fill_viridis(limits = c(0, 1),
                       "Maxent \nHabitat Suitability") +
  theme(axis.title.x = element_text(size=16),
        axis.title.y = element_text(size=16, angle=90),
        axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.key = element_blank())
```


## Plot ensemble
```{r}
# plot fancy map
# convert to a df for plotting in two steps,
# First, to a SpatialPointsDataFrame
wolf_EF_pts <- rasterToPoints(wolf_EnsembleResult_wm, spatial = TRUE)
# Then to a 'conventional' dataframe
wolf_EF_df  <- data.frame(wolf_EF_pts)
wolf_EF_df$HS <- wolf_EF_df$Wolf.2_EMwmeanByROC_mergedAlgo_mergedRun_mergedData/1000

wolf_EF_map <- ggplot(data = wolf_EF_df, aes(y=y, x=x)) +
  geom_raster(aes(fill=HS)) +
  theme_bw() +
  coord_equal() +
  scale_fill_viridis(limits = c(0, 1),
                       "Habitat Suitability") +
  theme(axis.title.x = element_text(size=16),
        axis.title.y = element_text(size=16, angle=90),
        axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.key = element_blank())
```


