---
title: "RegionalModels"
author: "Aimara Planillo"
date: "4/23/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Regional ensemble models

```{r}
source("./R/source_file.R")

workDir <- getwd()

dataproc_wd <- paste0(workDir, "/output/data_proc")
geoproc_wd <- paste0(workDir, "/output/geo_proc")
results_wd <- paste0(workDir, "/output/regional_ensemble")
```


## load data

```{r}
# telemetry data
gps_21ter2 <- st_read(paste0(dataproc_wd, "/gps_adults_21territories_allpoints_3035.gpkg")) %>% 
  st_transform(crs = 3035)

# mcp
mcp95_all_sf <- st_read(paste0(dataproc_wd, "/gps_MCP95_adults_allpoints_3035.gpkg")) %>% 
  st_transform(crs = 3035)

plot(st_geometry(mcp95_all_sf))

# wolf points for model
all_wolf_points <- st_read(paste0(dataproc_wd, "/all_wolf_points_ensemble_3035.gpkg")) %>% 
   st_transform(crs = 3035)
all_wolf_points$CLC2012[all_wolf_points$CLC2012 == 103] <- 102

str(all_wolf_points)
table(all_wolf_points$type, all_wolf_points$data)
```

# extract regions

we selected them previously manually, here we select the mcp and make a buffer around them to use as background data
```{r}
# give year to telemetry and mcp
gps_21ter2$year <- year(gps_21ter2$timestamp)

# Define regions
gps_df <- gps_21ter2 %>% 
  as.data.frame() %>% 
  dplyr::select(id,year)
table(gps_df)

# Region 1 = FT4, FT7, MT4, MT2, FT9, FT3, MT3, MT5, FT8, FT2
region1_id <- c("FT4", "FT7", "MT4", "MT2", "FT9", "FT3", "MT3", "MT5", "FT8", "FT2")
gps_df %>% 
  dplyr::filter(id %in% region1_id) %>% 
  group_by(year, id) %>% 
  summarise(n = n()) # from 2009 to 2015

# Region 2 = FT5, FT6, ID3, ID1, ID1p
region2_id <- c("FT5", "FT6", "ID3", "ID1", "ID1p")
gps_df %>% 
  dplyr::filter(id %in% region2_id) %>% 
  group_by(year, id) %>% 
  summarise(n = n()) # 2011 - 2018

# Region 3 = WR1, WR6, FT10, WF3
region3_id <- c("WR1", "WR6", "FT10", "WF3")
gps_df %>% 
  dplyr::filter(id %in% region3_id) %>% 
  group_by(year, id) %>% 
  summarise(n = n()) # all data from 2015 to 2018 


# get presences and bg points
wolf_presences <- all_wolf_points %>% 
  dplyr::filter(data == "presences")

wolf_bg <- all_wolf_points %>% 
  dplyr::filter(data == "background")

### extract region 1 data
# MCP
region1_mcp <- mcp95_all_sf %>%
  dplyr::filter(id %in% region1_id) %>% 
  st_cast("POLYGON")


# Presences within those mcps
region1_pres <- st_join(wolf_presences, region1_mcp, join = st_intersects) %>% 
  dplyr::filter(!is.na(id)) %>% 
  distinct(.keep_all = TRUE) %>% # remove duplicated coordinates due to overlapping mcps
  dplyr::select(-c(id, area, color))

# create buffer 50 km and dissolve
region1_buffer <- st_buffer(region1_mcp, 50000) %>% 
  st_union() %>% 
  st_sf() %>% 
  mutate(region = "region1") 

# extract bg points
n_pres <- nrow(region1_pres)
region1_bg <- st_join(wolf_bg, region1_buffer, join = st_within) %>% 
  sample_n(n_pres) %>%  # sample random rows number = n_pres
  dplyr::filter(!is.na(region)) %>% # remove points out of the polygon
  dplyr::select(-region)

plot(region1_buffer)
plot(st_geometry(region1_mcp), add = TRUE, col = "blue")
plot(st_geometry(region1_pres), add = TRUE, col = "green")
plot(st_geometry(region1_bg), add = TRUE, col = "red")

# merge data for analysis
region1_pres
region1_bg
region1_allpoints <- rbind(region1_pres, region1_bg)

```

## Run model region 1

```{r}
## Formating data
# Transform points into presence (1) - absence (0) and divide into train and test sets
region1_train <- region1_allpoints %>% 
  filter(fold != 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()
region1_train
table(region1_train$pres_bg)

nrow(region1_train) 

region1_test <- region1_allpoints %>% 
  filter(fold == 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()

region1_test
table(region1_test$pres_bg)

nrow(region1_test) 


## Extract environmental covariates
head(region1_allpoints)

region1_covariates_train <- region1_allpoints %>%
  filter(fold != 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(region1_covariates_train)
str(region1_covariates_train) 

region1_covariates_test <- region1_allpoints %>%
  filter(fold == 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(region1_covariates_test)
str(region1_covariates_test) 

region1_mod_dat <- BIOMOD_FormatingData(resp.var = region1_train,
                                     expl.var = region1_covariates_train,
                                   resp.name = "Wolf.region1")
region1_mod_dat
                                    
                                    
## Define models
# For maxent, we need to add some options to the models. 
# Important, the directory where to find java/maxent. 
myBiomodOptions <- BIOMOD_ModelingOptions(
  MAXENT.Phillips = list( path_to_maxent.jar = "C:/TEMP/maxent.jar",
                          maximumiterations = 5000,
                          responsecurves=TRUE,
                          jacknife=TRUE,
                          replicates=3,
                          visible = FALSE,
                          directory="/output/results/regional_ensemble/Wolf_region1",
                          linear = TRUE, 
                          quadratic = TRUE,
                          product = TRUE,
                          threshold = FALSE,
                          hinge = TRUE,
                          lq2lqptthreshold = 80,
                          l2lqthreshold = 10,
                          hingethreshold = 15,
                          beta_threshold = -1,
                          beta_categorical = -1,
                          beta_lqp = -1,
                          beta_hinge = -1,
                          defaultprevalence = 0.5))

region1_mod_out <- BIOMOD_Modeling(data = region1_mod_dat,
                            models =c('GLM','GAM','GBM', 'MARS', 'ANN','RF','MAXENT.Phillips'),
                            SaveObj = TRUE,
                           models.options = myBiomodOptions,
                           NbRunEval = 10,
                           DataSplit = 80, # common practice to validate: 80 train, 20 test
                           models.eval.meth = c('KAPPA','TSS','ROC'),
                           VarImport = 10)

region1_mod_out
str(region1_mod_out)
region1_mod_out@models.computed
# saveRDS(region1_mod_out, paste0(results_wd, "/region1_biomod_out.rds"))

# get all models evaluation
wolf_ModelEval <- get_evaluations(region1_mod_out)
wolf_ModelEval
# print the dimnames of this object
dimnames(wolf_ModelEval)

# get the AUC scores 
(AUC_table <- wolf_ModelEval["ROC","Testing.data",,,])
rowSums(AUC_table[,1:10])/10

# get the TSS scores 
(TSS_table <- wolf_ModelEval["TSS","Testing.data",,,])
rowSums(TSS_table[,1:10])/10

# write.csv(AUC_table, paste0(results_wd, "/region1_AUCTable.csv"))
# write.csv(TSS_table, paste0(results_wd, "/region1_TSSTable.csv"))
```

## get variable importance
```{r}
# print variable importances
wolf_VarImpor <- get_variables_importance(region1_mod_out)
wolf_VarImpor
# Variable importance of the full model
VarImpor_table <- wolf_VarImpor[,,"Full","AllData"]
VarImpor_table
# write.csv(VarImpor_table, paste0(results_wd,
#                                  "/region1_VariableImportance_FullModel.csv"))


## Average variable importance in the 10 model replicates
# extract the matrices to a list
wolf_VarImpor_list <- list()
for (i in 1:10){
  wolf_VarImpor_list[[i]] <- as.data.frame(wolf_VarImpor[,,i,])
}

for (i in 1:length(wolf_VarImpor_list)){
  wolf_VarImpor_list[[i]]$RUN <- paste0("Run_", i)
}

wolf_VarImport_allmodels <- do.call(rbind, wolf_VarImpor_list)
# write.csv(wolf_VarImport_allmodels, paste0(results_wd,
#                                            "/region1_VariableImportance_avg.csv"))
```


## Make ensemble
```{r}
### Building ensemble-models
# All models with AUC > 0.7
region1_ensemblemod <- BIOMOD_EnsembleModeling(
  modeling.output = region1_mod_out,
  chosen.models = 'all',
  em.by='all', 
  eval.metric = c('ROC'),
  eval.metric.quality.threshold = c(0.7), # model above 0.7 AUC values
  prob.mean.weight = T,
  prob.mean.weight.decay = 'proportional')

region1_ensemblemod
# saveRDS(region1_ensemblemod, paste0(results_wd, "/region1_ensemble_mod.rds"))

# get evaluation scores
get_evaluations(region1_ensemblemod)
```


## Project models
```{r}
# load rasters environmental rasters
env_germany <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_3035_AP.tif"))
names(env_germany) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads") 

env_germany2 <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_1km_3035_AP.tif"))
names(env_germany2) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads") 

### Project individual models
region1_BiomodProj <- BIOMOD_Projection(
  modeling.output = region1_mod_out,
  new.env = env_germany2,
  proj.name = 'region1',
  selected.models = 'all',
  compress = 'gzip',
  build.clamping.mask = F,
  output.format = '.grd', 
  do.stack=T)

region1_BiomodProj
# saveRDS(region1_BiomodProj, paste0(results_wd, "/region1_biomod_projection.rds"))


## Project ensemble-model
region1_BiomodEF <- BIOMOD_EnsembleForecasting(EM.output = region1_ensemblemod,
                            projection.output = NULL, 
                            new.env = env_germany2,
                            selected.models = 'all',
                            proj.name = "Region1",
                            compress = 'gzip')
region1_BiomodEF
# saveRDS(region1_BiomodEF, paste0(results_wd, "/region1_ensemble_forcasting.rds"))

region1_projEnsemble <- get_predictions(region1_BiomodEF)
names(region1_projEnsemble)

region1_EnsembleResult_wm <- region1_projEnsemble$Region1_EMwmeanByROC_mergedAlgo_mergedRun_mergedData # weighted mean model based on AUC

# writeRaster(region1_EnsembleResult_wm,
#             filename = paste0(results_wd, "/region1_ensemble_weightedmean_projection_3035.tif"),
#             overwrite = T)
```

## Calculate MESS for wolf predictions
```{r}
# Extract the environmental values in wolf presences
ref_env_r1 <- region1_allpoints %>% 
  as.data.frame() %>% 
  dplyr::filter(data == "presences") %>% 
  dplyr::select(CLC2012:Dist_roads)

# create the mess using full germany rasters
mess_wolf_r1 <- mess(env_germany, ref_env_r1, full = TRUE)
mess_wolf_r1 <- subset(mess_wolf_r1, 6, drop = FALSE)
# writeRaster(mess_wolf_r1, paste0(results_wd, "/region1_MESS_100m_3035.tif"), 
#             overwrite = T)
plot(mess_wolf_r1, breaks = c(-40, 40))
plot(st_geometry(region1_allpoints), add = TRUE)
```


# Region 2
```{r}
### extract region 2 data
# MCP
region2_mcp <- mcp95_all_sf %>%
  dplyr::filter(id %in% region2_id) %>% 
  st_cast("POLYGON")

# Presences within those mcps
region2_pres <- st_join(wolf_presences, region2_mcp, join = st_intersects) %>% 
  dplyr::filter(!is.na(id)) %>% 
  distinct(geom, .keep_all = TRUE) %>% 
  distinct(.keep_all = TRUE) %>% # remove duplicated coordinates due to overlapping mcps
  dplyr::select(-c(id, area, color))

# create buffer 50 km and dissolve
region2_buffer <- st_buffer(region2_mcp, 50000) %>% 
  st_union() %>% 
  st_sf() %>% 
  mutate(region = "region2") 

# extract bg points
n_pres <- nrow(region2_pres)
region2_bg <- st_join(wolf_bg, region2_buffer, join = st_within) %>% 
  dplyr::filter(!is.na(region)) %>% # remove points out of the polygon
  sample_n(n_pres) %>%  # sample random rows number = n_pres
  dplyr::select(-region)

plot(region2_buffer)
plot(st_geometry(region2_mcp), add = TRUE, col = "blue")
plot(st_geometry(region2_pres), add = TRUE, col = "green")
plot(st_geometry(region2_bg), add = TRUE, col = "red")

# merge data for analysis
region2_pres
region2_bg
region2_allpoints <- rbind(region2_pres, region2_bg)

st_write(region2_allpoints, paste0(dataproc_wd, 
                                   "/region2_allpoints_3035.gpkg"))
```

## Run model region 2

```{r}
## Formating data
# Transform points into presence (1) - absence (0) and divide into train and test sets
region2_train <- region2_allpoints %>% 
  filter(fold != 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()
region2_train
table(region2_train$pres_bg)

nrow(region2_train)

region2_test <- region2_allpoints %>% 
  filter(fold == 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()

region2_test
table(region2_test$pres_bg)

nrow(region2_test) 


## Extract environmental covariates
head(region2_allpoints)

region2_covariates_train <- region2_allpoints %>%
  filter(fold != 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(region2_covariates_train)
str(region2_covariates_train) 

region2_covariates_test <- region2_allpoints %>%
  filter(fold == 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(region2_covariates_test)
str(region2_covariates_test) 


region2_mod_dat <- BIOMOD_FormatingData(resp.var = region2_train,
                                     expl.var = region2_covariates_train,
                                    resp.name = "Wolf.region2")
region2_mod_dat
                                    
                                    
## Define models
# For maxent, we need to add some options to the models. 
# Important, the directory where to find java/maxent. 
myBiomodOptions <- BIOMOD_ModelingOptions(
  MAXENT.Phillips = list( path_to_maxent.jar = "C:/TEMP/maxent.jar",
                          maximumiterations = 5000,
                          responsecurves=TRUE,
                          jacknife=TRUE,
                          replicates=3,
                          visible = FALSE,
                          directory="/output/results/regional_ensemble/Wolf_region2",
                          linear = TRUE, 
                          quadratic = TRUE,
                          product = TRUE,
                          threshold = FALSE,
                          hinge = TRUE,
                          lq2lqptthreshold = 80,
                          l2lqthreshold = 10,
                          hingethreshold = 15,
                          beta_threshold = -1,
                          beta_categorical = -1,
                          beta_lqp = -1,
                          beta_hinge = -1,
                          defaultprevalence = 0.5))

region2_mod_out <- BIOMOD_Modeling(data = region2_mod_dat,
                            models =c('GLM','GAM','GBM', 'MARS', 'ANN','RF','MAXENT.Phillips'),
                            SaveObj = TRUE,
                           models.options = myBiomodOptions,
                           NbRunEval = 10,
                           DataSplit = 80, # common practice to validate: 80 train, 20 test
                           models.eval.meth = c('KAPPA','TSS','ROC'),
                           VarImport = 10)

region2_mod_out
str(region2_mod_out)
region2_mod_out@models.computed
# saveRDS(region2_mod_out, paste0(results_wd, "/region2_biomod_out.rds"))


# get all models evaluation
wolf_ModelEval <- get_evaluations(region2_mod_out)
wolf_ModelEval
# print the dimnames of this object
dimnames(wolf_ModelEval)

# get the AUC scores 
(AUC_table <- wolf_ModelEval["ROC","Testing.data",,,])
rowSums(AUC_table[,1:10])/10

# get the TSS scores 
(TSS_table <- wolf_ModelEval["TSS","Testing.data",,,])
rowSums(TSS_table[,1:10])/10

# write.csv(AUC_table, paste0(results_wd, "/region2_AUCTable.csv"))
# write.csv(TSS_table, paste0(results_wd, "/region2_TSSTable.csv"))
```

## get variable importance
```{r}
# print variable importances
wolf_VarImpor <- get_variables_importance(region2_mod_out)
wolf_VarImpor
# Variable importance of the full model
VarImpor_table <- wolf_VarImpor[,,"Full","AllData"]
VarImpor_table
# write.csv(VarImpor_table, paste0(results_wd,
#                                  "/region2_VariableImportance_FullModel.csv"))


## Average variable importance in the 10 model replicates
# extract the matrices to a list
wolf_VarImpor_list <- list()
for (i in 1:10){
  wolf_VarImpor_list[[i]] <- as.data.frame(wolf_VarImpor[,,i,])
}

for (i in 1:length(wolf_VarImpor_list)){
  wolf_VarImpor_list[[i]]$RUN <- paste0("Run_", i)
}

wolf_VarImport_allmodels <- do.call(rbind, wolf_VarImpor_list)
# write.csv(wolf_VarImport_allmodels, paste0(results_wd,
#                                            "/region2_VariableImportance_avg.csv"))
```


## Make ensemble
```{r}
### Building ensemble-models
# All models with AUC > 0.7
region2_ensemblemod <- BIOMOD_EnsembleModeling(
  modeling.output = region2_mod_out,
  chosen.models = 'all',
  em.by='all', 
  eval.metric = c('ROC'),
  eval.metric.quality.threshold = c(0.7), # model above 0.7 AUC values
  prob.mean = TRUE,
  prob.cv = FALSE,
  prob.ci = FALSE,
  prob.ci.alpha = 0.05,
  prob.median = FALSE,
  committee.averaging = FALSE,
  prob.mean.weight = TRUE,
  prob.mean.weight.decay = 'proportional')

region2_ensemblemod 
# saveRDS(region2_ensemblemod, paste0(results_wd, "/region2_ensemble_mod.rds"))
# get evaluation scores
get_evaluations(region2_ensemblemod)
```


## Project models
```{r}
# load rasters environmental rasters
env_germany <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_3035_AP.tif"))
names(env_germany) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads") 

# load aggregated rasters at 1 km resolution for faster projections
env_germany2 <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_1km_3035_AP.tif"))
names(env_germany2) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads")

### Project individual models
region2_BiomodProj <- BIOMOD_Projection(
  modeling.output = region2_mod_out,
  new.env = env_germany2,
  proj.name = 'region2',
  selected.models = 'all',
  compress = 'gzip',
  build.clamping.mask = F,
  output.format = '.grd', 
  do.stack=T)

region2_BiomodProj
# saveRDS(region2_BiomodProj, paste0(results_wd, "/region2_biomod_projection.rds"))

## Project ensemble-model
region2_BiomodEF <- BIOMOD_EnsembleForecasting(EM.output = region2_ensemblemod,
                            projection.output = region2_BiomodProj, 
                            selected.models = 'all',
                            proj.name = "Region2",
                            compress = 'gzip')
region2_BiomodEF
# saveRDS(region2_BiomodEF, paste0(results_wd, "/region2_ensemble_forcasting.rds"))

region2_projEnsemble <- get_predictions(region2_BiomodEF)
names(region2_projEnsemble)

region2_EnsembleResult_wm <- region2_projEnsemble$Wolf.region2_EMwmeanByROC_mergedAlgo_mergedRun_mergedData # weighted mean model
plot(region2_EnsembleResult_wm)

## It has gaps due to NAs in the means. We do it manually.
#### WEIGHTED MEAN
names(region2_indmodelProj)
length(names(region2_indmodelProj)) 

# get AUC values
region2_AUC_table <- read.csv(paste0(results_wd, "/region2_AUCTable.csv"), row.names = 1)
region2_AUC_values <- gather(region2_AUC_table)
region2_AUC_values <- region2_AUC_values %>%  
  mutate(model = paste0("Wolf.region2_AllData_", key, "_",rep(rownames(region2_AUC_table), 10)))
region2_AUC_values

# Remove models with AUC < 0.7
models_to_remove <- region2_AUC_values$model[region2_AUC_values$value < 0.7]
models_to_remove # 9
region2_indmodelProj_formean <- dropLayer(region2_indmodelProj, models_to_remove)
length(names(region2_indmodelProj_formean)) # 68

# compute weighted mean based on AUC values
region2_Results_wm <- calc(region2_indmodelProj,fun = weighted.mean, na.rm = TRUE, w = region2_AUC_values$value) 
plot(region2_Results_wm)

# writeRaster(region2_Results_wm,
#             filename = paste0(results_wd, "/region2_ensemble_weightedmean_projection_1km_3035.tif"),
#             overwrite = T)
```


## Calculate MESS for wolf predictions
```{r}
# Extract the environmental values in wolf presences
ref_env_r2 <- region2_allpoints %>% 
  as.data.frame() %>% 
  dplyr::filter(data == "presences") %>% 
  dplyr::select(CLC2012:Dist_roads)

# create the mess using full germany rasters
mess_wolf_r2 <- mess(env_germany, ref_env_r2, full = TRUE)
mess_wolf_r2 <- subset(mess_wolf_r2, 6, drop = FALSE)

# writeRaster(mess_wolf_r2, paste0(results_wd, "/region2_MESS_100m_3035.tif"), 
#             overwrite = T)
plot(mess_wolf_r2)
```



# Region 3

```{r}
### extract region 3 data
# MCP
region3_mcp <- mcp95_all_sf %>%
  dplyr::filter(id %in% region3_id) %>% 
  st_cast("POLYGON")


# Presences within those mcps
region3_pres <- st_join(wolf_presences, region3_mcp, join = st_intersects) %>% 
  dplyr::filter(!is.na(id)) %>% 
  distinct(.keep_all = TRUE) %>% # remove duplicated coordinates due to overlapping mcps
  dplyr::select(-c(id, area, color))

# create buffer 50 km and dissolve
region3_buffer <- st_buffer(region3_mcp, 50000) %>% 
  st_union() %>% 
  st_sf() %>% 
  mutate(region = "region3") 

# extract bg points
n_pres <- nrow(region3_pres)
region3_bg <- st_join(wolf_bg, region3_buffer, join = st_within) %>% 
  dplyr::filter(!is.na(region)) %>% # remove points out of the polygon
  sample_n(n_pres) %>%  # sample random rows number = n_pres
  dplyr::select(-region)

plot(region3_buffer)
plot(st_geometry(region3_mcp), add = TRUE, col = "blue")
plot(st_geometry(region3_pres), add = TRUE, col = "green")
plot(st_geometry(region3_bg), add = TRUE, col = "red")

# merge data for analysis
region3_pres
region3_bg
region3_allpoints <- rbind(region3_pres, region3_bg)

# st_write(region3_allpoints, paste0(dataproc_wd, 
#                                    "/region3_allpoints_3035.gpkg"))
```

## Run model region 3

```{r}
## Formating data
# Transform points into presence (1) - absence (0) and divide into train and test sets
region3_train <- region3_allpoints %>% 
  filter(fold != 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()
region3_train
table(region3_train$pres_bg)

nrow(region3_train) 

region3_test <- region3_allpoints %>% 
  filter(fold == 5) %>% 
  mutate(pres_bg = case_when(
    data == "presences" ~ 1,
    data == "background" ~ 0)) %>% 
  dplyr::select(pres_bg) %>% 
  as_Spatial()

region3_test
table(region3_test$pres_bg)

nrow(region3_test)

## Extract environmental covariates
head(region3_allpoints)

region3_covariates_train <- region3_allpoints %>%
  filter(fold != 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(region3_covariates_train)
str(region3_covariates_train) 

region3_covariates_test <- region3_allpoints %>%
  filter(fold == 5) %>% 
  as.data.frame() %>% 
  dplyr::select(CLC2012:Dist_roads) %>% 
  mutate(CLC2012 = as.factor(as.character(CLC2012)))
head(region3_covariates_test)
str(region3_covariates_test) 

region3_mod_dat <- BIOMOD_FormatingData(resp.var = region3_train,
                                     expl.var = region3_covariates_train,
                                    resp.name = "Region3")
region3_mod_dat
                                    
                                    
## Define models
# For maxent, we need to add some options to the models. 
# Important, the directory where to find java/maxent. 
myBiomodOptions <- BIOMOD_ModelingOptions(
  MAXENT.Phillips = list( path_to_maxent.jar = "C:/TEMP/maxent.jar",
                          maximumiterations = 5000,
                          responsecurves=TRUE,
                          jacknife=TRUE,
                          replicates=3,
                          visible = FALSE,
                          directory="/output/results/regional_ensemble/Wolf_region3/",
                          linear = TRUE, 
                          quadratic = TRUE,
                          product = TRUE,
                          threshold = FALSE,
                          hinge = TRUE,
                          lq2lqptthreshold = 80,
                          l2lqthreshold = 10,
                          hingethreshold = 15,
                          beta_threshold = -1,
                          beta_categorical = -1,
                          beta_lqp = -1,
                          beta_hinge = -1,
                          defaultprevalence = 0.5))

region3_mod_out <- BIOMOD_Modeling(data = region3_mod_dat,
                            models =c('GLM','GAM','GBM', 'MARS', 'ANN','RF','MAXENT.Phillips'),
                            SaveObj = TRUE,
                           models.options = myBiomodOptions,
                           NbRunEval = 10,
                           DataSplit = 80, # common practice to validate: 80 train, 20 test
                           models.eval.meth = c('KAPPA','TSS','ROC'),
                           VarImport = 10)

region3_mod_out
str(region3_mod_out)
region3_mod_out@models.computed
# saveRDS(region3_mod_out, paste0(results_wd, "/region3_biomod_out.rds"))

# get all models evaluation
region3_ModelEval <- get_evaluations(region3_mod_out)
region3_ModelEval
# print the dimnames of this object
dimnames(region3_ModelEval)

# get the AUC scores 
(AUC_table <- region3_ModelEval["ROC","Testing.data",,,])
rowSums(AUC_table[,1:10])/10

# get the TSS scores 
(TSS_table <- region3_ModelEval["TSS","Testing.data",,,])
rowSums(TSS_table[,1:10])/10

# write.csv(AUC_table, paste0(results_wd, "/region3_AUCTable.csv"))
# write.csv(TSS_table, paste0(results_wd, "/region3_TSSTable.csv"))
```

## get variable importance
```{r}
# print variable importances
region3_VarImpor <- get_variables_importance(region3_mod_out)
region3_VarImpor
# Variable importance of the full model
VarImpor_table <- region3_VarImpor[,,"Full","AllData"]
VarImpor_table
# write.csv(VarImpor_table, paste0(results_wd,
#                                  "/region3_VariableImportance_FullModel.csv"))


## Average variable importance in the 10 model replicates
# extract the matrices to a list
region3_VarImpor_list <- list()
for (i in 1:10){
  region3_VarImpor_list[[i]] <- as.data.frame(region3_VarImpor[,,i,])
}

for (i in 1:length(region3_VarImpor_list)){
  region3_VarImpor_list[[i]]$RUN <- paste0("Run_", i)
}

region3_VarImport_allmodels <- do.call(rbind, region3_VarImpor_list)
# write.csv(region3_VarImport_allmodels, paste0(results_wd,                                           "/region3_VariableImportance_avg.csv"))
```


## Make ensemble
```{r}
### Building ensemble-models
# All models with AUC > 0.7

get_built_models(region3_mod_out)
region3_ensemblemod <- BIOMOD_EnsembleModeling(
  modeling.output = region3_mod_out,
  # chosen.models = 'all',
  chosen.models = get_built_models(region3_mod_out),
  em.by='all', 
  eval.metric = c('ROC'),
  eval.metric.quality.threshold = c(0.7), # model above 0.7 AUC values
  prob.mean = T,
  prob.cv = T,
  prob.ci = T,
  prob.ci.alpha = 0.05,
  prob.median = T,
  committee.averaging = T,
  prob.mean.weight = T,
  prob.mean.weight.decay = 'proportional')

region3_ensemblemod
# saveRDS(region3_ensemblemod, paste0(results_wd, "/region3_ensemble_mod.rds"))
# get evaluation scores
get_evaluations(region3_ensemblemod)
```


```{r}
region3_ensemblemod <- BIOMOD_EnsembleModeling( 
                 modeling.output = region3_mod_out,
                 chosen.models = get_built_models(region3_mod_out), 
                 em.by = 'all',
                 eval.metric = c('ROC'),
                 eval.metric.quality.threshold = c(0.7),
                 prob.mean = FALSE,
                 prob.cv = FALSE,
                 prob.ci = FALSE,
                 prob.ci.alpha = 0.05,
                 prob.median = FALSE,
                 committee.averaging = TRUE,
                 prob.mean.weight = FALSE,
                 prob.mean.weight.decay = 'proportional' )



test <- get_predictions(region3_ensemblemod)
region3_indmodelProj
region3_Result <- calc(region3_indmodelProj,fun = mean, na.rm = TRUE); #Choose whatever descriptive statistic you'd like
region3_Result <- calc(region3_indmodelProj[[1]],fun = mean) #Choose whatever descriptive statistic you'd like
plot(region3_Result)
region3_Result <- calc(region3_indmodelProj[[11:20]],fun = mean) #Choose whatever descriptive statistic you'd like
plot(region3_Result)
```


## Project models
```{r}
# load rasters environmental rasters
env_germany <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_3035_AP.tif"))
names(env_germany) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads") 

# aggregate rasters at 1 km resolution for faster projections
env_germany2 <- aggregate(env_germany, fact = 10, fun = mean, na.rm = TRUE)
# writeRaster(env_germany2, paste0(geoproc_wd, "/stack_env_rasters_germany_1km_3035_AP.tif"))
env_germany2 <- stack(env_germany2)

### Project individual models
region3_BiomodProj <- BIOMOD_Projection(
  modeling.output = region3_mod_out,
  new.env = env_germany2,
  proj.name = 'region3',
  selected.models = get_built_models(region3_mod_out),
  compress = 'gzip',
  build.clamping.mask = F,
  output.format = '.grd', 
  do.stack=T, 
  na.omit = FALSE)

region3_mod_out@models.computed
region3_BiomodProj
# saveRDS(region3_BiomodProj, paste0(results_wd, "/region3_biomod_projection.rds"))


## Project ensemble-model
region3_BiomodEF <- BIOMOD_EnsembleForecasting(EM.output = region3_ensemblemod,
                            projection.output = region3_BiomodProj,
                           proj.name = "Region3",
                            compress = 'gzip')
region3_BiomodEF
# saveRDS(region3_BiomodEF, paste0(results_wd, "/region3_ensemble_forcasting.rds"))

region3_projEnsemble <- get_predictions(region3_BiomodEF)
names(region3_projEnsemble)

region3_EnsembleResult_wm <- region3_projEnsemble$Region3_EMwmeanByROC_mergedAlgo_mergedRun_mergedData  # weighted mean model based on AUC
plot(region3_EnsembleResult_wm)
# writeRaster(region3_EnsembleResult_wm,
#             filename = paste0(results_wd, "/region3_ensemble_weightedmean_projection_3035.tif"),
#             overwrite = T)
```


## Calculate MESS for wolf predictions
```{r}
# Extract the environmental values in wolf presences
ref_env_r3 <- region3_allpoints %>% 
  as.data.frame() %>% 
  dplyr::filter(data == "presences") %>% 
  dplyr::select(CLC2012:Dist_roads)

# create the mess using full germany rasters
mess_wolf_r3 <- mess(env_germany, ref_env_r3, full = TRUE)
mess_wolf_r3 <- subset(mess_wolf_r3, 6, drop = FALSE)
mess_wolf_r3
# writeRaster(mess_wolf_r3, paste0(results_wd, "/region3_MESS_100m_3035.tif"), 
#             overwrite = T);
plot(mess_wolf_r3)
```




## Sizes of the regions
```{r}
region1_mcp
region2_mcp
region3_mcp

plot(st_union(region1_mcp))
plot(st_union(region2_mcp))
plot(st_union(region3_mcp))

st_area(st_union(region1_mcp))

st_area(st_union(region2_mcp))

st_area(st_union(region3_mcp))
```

