---
title: "5_N_territory_estimation"
author: "Aimara Planillo"
date: "15/04/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# libraries and functions
source("./R/source_file.R")
source("./R/source_wd.R")

workDir <- getwd()

dataproc_wd <- paste0(workDir, "/output/data_proc")
geoproc_wd <- paste0(workDir, "/output/geo_proc")
results_wd <- paste0(workDir, "/output/results")
```


# Territory estimation in Germany

We are going to create a map at 10 km resolution to estimate the total number of 
wolf territories that potentially could happen in Germany (max number)  

We need to rescale the 100 m resolution predictions to 10 km, using Eurogrid as 
a template for the grid cells at 10 km resolution.

we are going to use the mean value of the new cells and the sd to create a binary map
with the potential territories

### Load data
```{r}
# ensemble 100 m
wolf_EnsembleResult_wm <- raster(paste0(results_wd, "/ensemble_global/ensemble_samebgpres_weightedmean_projection_3035.tif"))

# template Eurogrid
eurogrid_10km <- st_read(paste0(geoproc_wd, "/DE_10km_final_Union/DE_10km_final_Union1.shp")) %>% 
  st_transform(crs = 3035)
plot(st_geometry(eurogrid_10km))


# Environmental values for germany
env_germany <- stack(paste0(geoproc_wd, "/stack_env_rasters_germany_3035_AP.tif"))
names(env_germany) <- c("CLC2012", "Pop_den", "HFP", "Dist_settl", "Dist_roads") 

# Water mask to remove territories on water from env. layers
watermask_10km <- raster(paste0(geoproc_wd, "/watermask_10km_3035.tif"))

# load germany border
germany_3035 <- read_sf(paste0(geoproc_wd, "/germany_3035.gpkg"), crs = 3035) 

# territory data
# load all GPS data for resident adults
terr_sf <- st_read(paste0(dataproc_wd, "/territory_points_3035.gpkg")) %>% 
  st_transform(crs = 3035)
terr_sf
```


## Resample

We are going to use the Eurogrid as a template for the 10 km grid cells
We also build two models at the 10 km resolution:
- Mean value
- Sd value

```{r}
# create new raster from grid at same resolution than predictions
Eurogrid_raster_100m <- raster(extent(eurogrid_10km), res = 100, crs = crs(as(eurogrid_10km,"Spatial")))
values(Eurogrid_raster_100m) <- 1

# give values of predictions to new raster
resampled_raster_100m <- resample(x = wolf_EnsembleResult_wm, 
                                  y = Eurogrid_raster_100m, 
                                  method="bilinear" )

# get mean values at 10km resolution
pred_mean_10km <- aggregate(resampled_raster_100m, fact = 100, fun = mean, na.rm = TRUE)
# writeRaster(pred_mean_10km, paste0(results_wd, "/10km_res/ensemble_mean_10km_3035.tif"), overwrite = TRUE)

# get sd values at 10km resolution
pred_sd_10km <- aggregate(resampled_raster_100m, fact = 100, fun = sd, na.rm = TRUE)
# writeRaster(pred_sd_10km, paste0(results_wd, "/10km_res/ensemble_sd_10km_3035.tif"), overwrite = TRUE)

## Extract data for territory centroids
terr_mean <- raster::extract(pred_mean_10km, as_Spatial(terr_sf), sp = TRUE)
terr_sd <- raster::extract(pred_sd_10km, terr_mean, sp = TRUE)
names(terr_sd)[6] <- "ensemble_mean"
names(terr_sd)[7] <- "ensemble_sd"

terr_ensemble_values_10km <- st_as_sf(terr_sd)
# st_write(terr_ensemble_values_10km, paste0(results_wd, "/10km_res/terr_ensemble_values_10km_3035.gpkg"),
#          append=FALSE)

plot(terr_ensemble_values_10km)

# Create data frame to explore the values
terr_values_df <- as.data.frame(terr_ensemble_values_10km)

hs_terr_plot <- ggplot(terr_values_df, aes(x = ensemble_mean, y = ensemble_sd)) +
  geom_smooth() +
  geom_point() +
  ggtitle("Mean and sd of Habitat suitability values \nin wolf territories") +
  theme(
    panel.background = element_blank(),
    axis.line = element_line("black"), 
    plot.title = element_text(color = "black", size = 16)
  )
```

# Create binary map

## Mean - sd

```{r}
# count each territory only once for the average calculations. Remove duplicated entries
unique_terr <- terr_data %>%
  dplyr::select(Territorium, Tkurz, ensemble_mean, ensemble_sd) %>% 
  distinct(.keep_all = TRUE) %>%
  as.data.frame()

mean_hs <- mean(unique_terr$ensemble_mean)
mean_hs
sd_hs <- sd(unique_terr$ensemble_mean)
sd_hs 

min_hs <- min(unique_terr$ensemble_mean)
max_hs <- max(unique_terr$ensemble_mean)
min_hs 
max_hs 

# Binary mean - sd
my_threshold <- mean_hs - sd_hs

binary_tmp <- pred_mean_10km
binary_tmp[binary_tmp < my_threshold] <- 0
binary_tmp[binary_tmp >= my_threshold] <- 1

binary_1 <- binary_tmp


# Binary low 95% CI
my_threshold <- mean_hs - 1.96*sd_hs 

binary_tmp <- pred_mean_10km
binary_tmp[binary_tmp < my_threshold] <- 0
binary_tmp[binary_tmp >= my_threshold] <- 1

binary_2 <- binary_tmp


plot(binary_1)
plot(binary_2)
```


# Create binary map

## TSS
```{r}
# get evaluation scores and maxTSS threshold
scores_all <- get_evaluations(wolf_ensemblemod)
threshold <- scores_all$Wolf.2_EMwmeanByROC_mergedAlgo_mergedRun_mergedData["TSS", "Cutoff"]
threshold

# Binary maxTSS threshold

binary_tmp <- pred_mean_10km
binary_tmp[binary_tmp < threshold] <- 0
binary_tmp[binary_tmp >= threshold] <- 1

binary_maxTSS <- binary_tmp

plot(binary_maxTSS)
```


# Territory count

## Remove cell that are alone (not territories) and water
```{r}
## Binary 1: Mean - sd 
my_focal_map_terr <- binary_1 * watermask_10km
plot(my_focal_map_terr)

## compute focal mean to identfy all raster cells having a neighbour
my_focal_map_terr_sum <- raster::focal(x = my_focal_map_terr, 
                                  w = matrix(c(1, 1, 1,
                                               1, 1, 1,
                                               1, 1, 1),
                                             nrow=3, 
                                             ncol=3),  
                                  fun = sum, 
                                  na.rm = TRUE, 
                                  pad = TRUE, 
                                  padValue = 1
                                  )

final_map_terr <- my_focal_map_terr_sum
final_map_terr[] <- 0
final_map_terr[my_focal_map_terr_sum > 1 & my_focal_map_terr == 1 ] <- 1
plot(final_map_terr)

# writeRaster(final_map_terr, 
#             paste0(results_wd, "/10km_res/ensemble_binary_mean-sd_watermask_10km_3035.tif"))


plot(mask(final_map_terr, germany_3035))
plot(mcps_plot, add =T, alpha = 0.5)
# Number of territories
floor(sum(values(final_map_terr))/2)

##Binary 2: low 95% CI
my_focal_map_terr <- binary_2 * watermask_10km
plot(my_focal_map_terr)

## compute focal mean to identfy all raster cells having a neighbour
my_focal_map_terr_sum <- raster::focal(x = my_focal_map_terr, 
                                  w = matrix(c(1, 1, 1,
                                               1, 1, 1,
                                               1, 1, 1),
                                             nrow=3, 
                                             ncol=3),  
                                  fun = sum, 
                                  na.rm = TRUE, 
                                  pad = TRUE, 
                                  padValue = 1
                                  )

final_map_terr <- my_focal_map_terr_sum
final_map_terr[] <- 0
final_map_terr[my_focal_map_terr_sum > 1 & my_focal_map_terr == 1 ] <- 1
plot(final_map_terr)

# writeRaster(final_map_terr, 
#             paste0(results_wd, "/10km_res/ensemble_binary_low95CI_watermask_10km_3035.tif"))


final_map_terr_plot <- mask(final_map_terr, germany_3035)

plot(mask(final_map_terr, germany_3035))
plot(mcps_plot, add =T, alpha = 0.5)
# Number of territories
floor(sum(values(final_map_terr))/2)
```




# Analysis of Reproduction in stablished territories
```{r}
## load data
terr_ensemble_values_10km <- st_read(paste0(results_wd,
                                            "/10km_res/terr_ensemble_values_10km_3035.gpkg")) %>% 
  st_transform(crs = 3035)

plot(terr_ensemble_values_10km)

summary(terr_ensemble_values_10km)


# Create data to plot histograms of reproductive events 
terr_values_df_rep <- terr_ensemble_values_10km %>% 
  as.data.frame() %>% 
  mutate(reproduction = case_when(
    Welpen > 0 ~ 1, 
    Welpen == 0 ~ 0)) %>%
  group_by(Territorium) %>% 
  mutate(terr_age = row_number()) %>% 
  ungroup() %>% 
  as.data.frame()


head(terr_values_df_rep, 30)  
terr_values_df_rep[terr_values_df_rep$Welpen == 13,]

# Model with all effects
library(lme4)

mod_rep <- glmer(reproduction ~ scale(ensemble_mean) + scale(ensemble_sd) + scale(terr_age) + (1|Territorium), 
                  family = "binomial", 
                  data = terr_values_df_rep)
summary(mod_rep)

plot(ggeffect(mod_rep, terms = "ensemble_mean"))
plot(ggeffect(mod_rep, terms = "ensemble_sd"))
plot(ggeffect(mod_rep, terms = "terr_age"))
saveRDS(mod_rep, paste0(results_wd, "/10km_res/model_reproduction_terr_habitatsuitability.rds"))


# Model with mean 
mod_mean <- glmer(reproduction ~ scale(ensemble_mean) + scale(terr_age) + (1|Territorium), 
                  family = "binomial", 
                  data = terr_values_df_rep, 
                  glmerControl(optimizer = "bobyqa"))
summary(mod_mean)
plot(ggeffect(mod_mean, terms = "ensemble_mean"))

# Model with sd 
mod_sd <- glmer(reproduction ~ scale(ensemble_sd) + scale(terr_age) + (1|Territorium), 
                  family = "binomial", 
                  data = terr_values_df_rep)
summary(mod_sd)
plot(ggeffect(mod_sd, terms = "ensemble_sd"))


## With number of offspring
mod_repnumber <- glmer(Welpen ~ scale(ensemble_mean) + scale(ensemble_sd) + scale(terr_age) + (1|Territorium), 
                  family = "poisson", 
                  data = terr_values_df_rep)
summary(mod_repnumber)

plot(ggeffect(mod_repnumber, terms = "ensemble_sd"))
plot(ggeffect(mod_repnumber, terms = "ensemble_mean"))

```

